\documentclass[11pt,a4paper]{article}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage{amssymb}
\usepackage[margin=24mm]{geometry}
\usepackage{xcolor}
\usepackage{microtype}
\DisableLigatures{encoding=T1,family=tt*}
\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{longtable}
\usepackage{listings}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,shapes.geometric}
\usepackage{fancyhdr}
\usepackage{hyperref}

\definecolor{ZipperInk}{HTML}{17212B}
\definecolor{ZipperBlue}{HTML}{176B87}
\definecolor{ZipperLight}{HTML}{EEF6F8}
\definecolor{CodeBackground}{HTML}{F5F7F8}
\definecolor{CodeComment}{HTML}{58715D}
\definecolor{CodeKeyword}{HTML}{7A3E9D}
\definecolor{WarningBackground}{HTML}{FFF6DF}
\definecolor{WarningBorder}{HTML}{B7791F}
\definecolor{ZipperGreen}{HTML}{2F855A}
\definecolor{ZipperGold}{HTML}{D69E2E}
\definecolor{ZipperRed}{HTML}{C05640}

\hypersetup{
  colorlinks=true,
  linkcolor=ZipperBlue,
  urlcolor=ZipperBlue,
  citecolor=ZipperBlue,
  pdftitle={ZipperGen Workflow Development and Deployment},
  pdfauthor={The ZipperGen Authors}
}

\pagestyle{fancy}
\fancyhf{}
\lhead{ZipperGen development and deployment}
\rhead{Beginner's end-to-end guide}
\cfoot{\thepage}
\setlength{\headheight}{14pt}

\setlist[itemize]{topsep=4pt,itemsep=2pt}
\setlist[enumerate]{topsep=4pt,itemsep=3pt}
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt}

\lstset{
  basicstyle=\ttfamily\small,
  backgroundcolor=\color{CodeBackground},
  frame=single,
  rulecolor=\color{black!18},
  breaklines=true,
  breakatwhitespace=false,
  columns=fullflexible,
  keepspaces=true,
  showstringspaces=false,
  upquote=true,
  keywordstyle=\color{CodeKeyword}\bfseries,
  commentstyle=\color{CodeComment},
  stringstyle=\color{ZipperBlue},
  numbers=left,
  numberstyle=\tiny\color{black!45},
  numbersep=8pt,
  xleftmargin=2em,
  framexleftmargin=1.5em
}

\lstdefinestyle{shell}{
  language=bash,
  numbers=none
}

\lstdefinestyle{python}{
  language=Python
}

\newcommand{\menu}[1]{\textbf{#1}}
\newcommand{\placeholder}[1]{\textit{$\langle$#1$\rangle$}}

\tikzset{
  zgstage/.style={
    draw=ZipperBlue,
    fill=ZipperLight,
    very thick,
    rounded corners=2.5pt,
    align=center,
    minimum height=12mm,
    inner sep=4pt,
    font=\small
  },
  zgplain/.style={
    draw=ZipperInk!55,
    fill=white,
    thick,
    rounded corners=2.5pt,
    align=center,
    minimum height=11mm,
    inner sep=4pt,
    font=\small
  },
  zgstore/.style={
    cylinder,
    shape border rotate=90,
    aspect=0.22,
    draw=ZipperGreen,
    fill=ZipperGreen!10,
    very thick,
    align=center,
    minimum height=16mm,
    minimum width=30mm,
    inner sep=4pt,
    font=\small
  },
  zgflow/.style={-{Latex[length=2.2mm,width=1.5mm]}, thick, draw=ZipperBlue},
  zgdashed/.style={-{Latex[length=2.2mm,width=1.5mm]}, thick, dashed,
    draw=ZipperRed},
  zglabel/.style={font=\scriptsize, align=center, fill=white, inner sep=1.5pt}
}

\newenvironment{notebox}[1]
{\begin{center}\begin{minipage}{0.94\linewidth}
 \color{ZipperInk}\hrule\vspace{5pt}\textbf{#1}\par\smallskip}
{\vspace{5pt}\hrule\end{minipage}\end{center}}

\newenvironment{warningbox}[1]
{\begin{center}\begin{minipage}{0.94\linewidth}
 \color{WarningBorder}\hrule\vspace{5pt}
 \color{ZipperInk}\textbf{#1}\par\smallskip}
{\vspace{5pt}\color{WarningBorder}\hrule\end{minipage}\end{center}}

\title{\color{ZipperInk}\Huge\textbf{ZipperGen Workflow Development and Deployment}\\[5pt]
\Large A complete prompt-to-production guide for beginning users}
\author{The ZipperGen Authors}
\date{Guide edition: 22 July 2026}

\begin{document}

\maketitle

\begin{abstract}
This guide walks through one complete ZipperGen development cycle. It begins
with ZipperGen Studio, the recommended one-terminal path for selecting,
maintaining one canonical specification, inspecting project/workflow state, assigning
models and providers, running, resuming, refining, and deploying a workflow
without managing environment exports, SQLite filenames, or human-task
identifiers.
The remainder then explains every underlying command and artifact in detail:
installation, code views, durable approval, coding-assistant refinement,
semantic comparison, named deployment, logs, diagnosis, restart, and stop. It
is written for readers who can use a terminal and editor but do not need to be
experienced Python or operations engineers.
\end{abstract}

\begin{notebox}{The main promise}
The workflow is always ordinary Python code. A coding assistant may create or
change that code from one visible, versionable workflow specification, but ZipperGen
performs the deterministic work: loading it, validating every projection,
showing what each participant does, comparing semantics, storing durable
execution state, and managing deployment. LLMs propose workflows; ZipperGen
makes them explicit, verifiable, and deliberately deployable. It never treats
opaque model output as an automatically deployed workflow.
\end{notebox}

\tableofcontents
\newpage

\section{Start here: the ZipperGen Studio path}
\label{sec:studio}

ZipperGen Studio is the recommended way to experience the development cycle.
It is a lightweight, project-aware command environment inside the ordinary
terminal. It does not replace the operating-system shell, Git, the editor, or
the visible Python workflow. Instead, it remembers the project context and
turns the most common operations into short, discoverable commands.

This opening section is self-contained: it creates a user-owned workflow
project and clones the ZipperGen framework below it. The project root contains
the canonical specification, generated workflows, tests, and its own Git history; the nested
\path{zippergen} directory contains the framework checkout. Readers who already
have both may skip only that first subsection. Section~\ref{sec:install} later
explains prerequisites and alternatives in more detail.

\begin{notebox}{What is required, and what is optional}
The free local path requires macOS or Linux, a terminal, an Internet connection
for the initial clone and installation, and enough familiarity to copy commands
and edit a file. The built-in mock LLM needs no API key and makes no paid model
calls. Codex or another repository-aware coding assistant is needed only to
generate or refine source from a prompt; the shipped example is the fallback.
OpenAI and Anthropic accounts, real API keys, and supervised service startup are
all optional exercises, clearly labelled below.
\end{notebox}

\begin{notebox}{What Studio removes from the normal development path}
For the guided local cycle, you do not need to export \texttt{ZIPPERGEN\_HOME},
invent a SQLite filename, copy a task ID, keep two terminals synchronized, or
remember the inspection flags. Studio creates unique durable stores, presents
human tasks inline, and remembers the current workflow, run, view, and named
deployment for this project.
\end{notebox}

\subsection{Create the tutorial checkout and local environment}

Open an ordinary terminal. Create the tutorial project, initialize its Git
history, clone the framework below it, and install the framework's command as
an editable user tool:

\begin{lstlisting}[style=shell]
mkdir -p "$HOME/Projects"
cd "$HOME/Projects"
mkdir zippergen-tutorial
cd "$HOME/Projects/zippergen-tutorial"
git init
git clone https://github.com/zippergen-io/zippergen.git zippergen
uv python install 3.12
uv sync --project zippergen --python 3.12
uv tool install --force --editable ./zippergen
zippergen --help
\end{lstlisting}

The parent is now the tutorial project and coding-assistant root. Its nested
\path{zippergen} directory is a separate framework checkout containing
\path{AGENTS.md}, the workflow skill, framework source, and shipped examples.
Studio will create the application project's fixed \path{specification.md}
when you enter \texttt{create}; there is no prompt directory or prompt filename
to choose. The ignored \path{.zippergen/pending-refinement.md} is likewise
created automatically only when a change is pending.
\texttt{uv sync} creates the framework's local environment;
\texttt{uv tool install --force --editable} makes the short \texttt{zippergen} command
available while still using this editable checkout. The project, checkout,
downloaded Python, and environment survive closing the terminal or rebooting.
The final help output should list \texttt{studio} among the available
top-level commands. You do not run \texttt{models} here: \texttt{models} is a
short command entered later inside Studio.

\begin{notebox}{If your existing tutorial checkout is flat}
An older version of this guide cloned ZipperGen directly into
\path{zippergen-tutorial}. In that layout the project root itself contains
\path{AGENTS.md}, \path{pyproject.toml}, and \path{src}, and
there is no nested \path{zippergen} directory. Do not add a
\texttt{zippergen/} prefix to framework paths. The Studio creation command is
still simply \texttt{create}. For that flat checkout, run these equivalent
installation commands from its root:
\begin{quote}
\texttt{uv sync --python 3.12}\\
\texttt{uv tool install --force --editable .}
\end{quote}
\end{notebox}

If \texttt{git} is not found, install the standard Git package for your
operating system and repeat the first listing. If \texttt{uv} is not found but
\texttt{curl} is available, install it with:

\begin{lstlisting}[style=shell]
curl -LsSf https://astral.sh/uv/install.sh | sh
\end{lstlisting}

Restart the terminal if the installer requests it, then continue from:

\begin{lstlisting}[style=shell]
cd "$HOME/Projects/zippergen-tutorial"
uv python install 3.12
uv sync --project zippergen --python 3.12
uv tool install --force --editable ./zippergen
zippergen --help
\end{lstlisting}

If \path{$HOME/Projects/zippergen-tutorial/zippergen} already contains the
framework checkout, do not clone over it. Enter the project and inspect both
Git repositories separately:

\begin{lstlisting}[style=shell]
cd "$HOME/Projects/zippergen-tutorial"
git status --short
git -C zippergen status --short
\end{lstlisting}

An empty result from each command means that repository has no uncommitted
changes. Before the first Studio \texttt{project init}, the outer command may
show \texttt{?? zippergen/}; project initialization adds the local framework
checkout to the outer \path{.gitignore}. Do not run
\texttt{git pull}, delete files, or reset changes you do not understand merely
to begin the tutorial.

\subsection{The short path and the transparent path}

Studio and the detailed CLI operate on the same Python workflows, semantic
views, SQLite runtime, and guided deployer. Studio is therefore not a second
execution model or a visual builder. The later chapters remain important when
you want automation, CI, an explicit store, an external approval adapter, or a
precise understanding of every generated artifact.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[node distance=5.5mm]
  \node[zgstage, text width=23mm] (intent) {Natural-language\\intent};
  \node[zgstage, text width=24mm, right=of intent] (brief) {Studio task from\\specification};
  \node[zgstage, text width=23mm, right=of brief] (assistant) {Repository-aware\\coding assistant};
  \node[zgstage, text width=23mm, right=of assistant] (source) {Visible Python\\workflow};
  \node[zgstage, text width=24mm, right=of source] (studio) {Show, validate,\\run, deploy};
  \node[zgstore, text width=31mm, below=10mm of brief] (ledger)
    {Canonical intent\\\texttt{specification.md}};
  \node[zgstore, text width=31mm, below=10mm of studio] (state)
    {Managed project state\\\texttt{\$HOME/.zippergen}};
  \draw[zgflow] (intent) -- (ledger);
  \draw[zgflow] (ledger) -- (brief);
  \draw[zgflow] (brief) -- (assistant);
  \draw[zgflow] (assistant) -- (source);
  \draw[zgflow] (source) -- (studio);
  \draw[zgdashed] (studio) -- (state);
\end{tikzpicture}
\caption{Studio turns natural-language intent into one versioned project
specification before preparing a coding-assistant handoff. Visible Python is
then inspected by deterministic ZipperGen operations; runs and secrets remain
outside the Git checkout.}
\label{fig:studio-path}
\end{figure}

\subsection{The complete prompt-to-deployment walk-through}

This subsection carries out every box in Figure~\ref{fig:studio-path} in small
steps. It creates a reviewed-answer workflow from a natural-language
specification. Studio and the coding assistant handle its Python filename and
symbol. The shipped \path{zippergen/examples/tutorial_review.py:tutorial_review}
is the deterministic fallback if a coding assistant is not available.

\subsubsection*{Step 1: open Studio without selecting a workflow}

Keep the first terminal in the Git root and start Studio:

\begin{lstlisting}[style=shell]
cd "$HOME/Projects/zippergen-tutorial"
zippergen
\end{lstlisting}

Expected beginning:

\begin{lstlisting}[style=shell]
ZipperGen Studio
Project: /path/to/zippergen-tutorial
No workflow selected.
Type a command or describe what you want in ordinary language;
press Tab to complete; 'help' shows the exact commands.
zippergen [no workflow]>
\end{lstlisting}

Everything after the last line is a Studio command. Do not prefix Studio
commands with \texttt{zippergen}. Enter \texttt{help} now; it is the
built-in reminder for creation, selection, inspection, model routing,
execution, refinement, deployment, and operation.

Press Tab at any point while entering a command. Studio completes commands and
subcommands and, where relevant, offers the workflows, participants,
LLM-active participants, providers, remembered deployment, and project files
that are valid in the current context. For example, type \texttt{sp} followed
by Tab to obtain \texttt{spec}; after selecting the tutorial workflow,
\texttt{show agent W} followed by Tab completes \texttt{Writer}. When several
matches exist, the completion menu explains each one. When exactly one match
exists---for example, \texttt{resu} for \texttt{resume}---the bottom line shows
that match and its explanation even though no menu is necessary. Use the up
and down arrows for private per-project command history. A faint suggestion
from that history is accepted with the right arrow. This interactive support
is provided by \texttt{prompt-toolkit}; scripts and piped input keep the plain
non-interactive input path.

Initialize the visible project contract, then inspect the initial dashboard:

\begin{lstlisting}[style=shell]
project init zippergen-tutorial
current
\end{lstlisting}

Studio creates \path{zippergen.toml} and safe Git
ignores for \path{.zippergen}, the nested framework checkout, and the optional
transparent runtime. The manifest and the \path{specification.md} created in
Step~2 are project source and may be committed. Timestamped tasks, semantic baselines, runs, and secrets remain
in private Studio state outside the project. Studio mirrors only the current
generated task at \path{.zippergen/current-task.md}; Git ignores that file.
Repeating \texttt{project init} is safe; it reports the existing manifest
instead of overwriting it.

If you later need a reset, do not delete hidden directories or visible project
files by hand. Plain \texttt{project reset} first asks which scope you mean:
\emph{Fresh design cycle}, \emph{Studio state only}, or \emph{Cancel}. The
state-only choice archives runs and stores, assistant-task/refinement/command
history, model/provider preferences, development secrets, generated tasks, and
pending refinement while retaining every visible project file. The fresh
choice additionally archives \path{zippergen.toml},
\path{specification.md}, and legacy prompt material. Both keep workflow source,
tests, Git history, the nested framework checkout, and deployments. Archives
are owner-only below \path{ZIPPERGEN_HOME/resets}.

This distinction explains the complete restart sequence. Choose the fresh
design cycle, confirm it, then enter \texttt{project init} and \texttt{create}.
The manifest is genuinely new and the specification contains only the writing
guide described in Step~2, not the previous requirements. For automation, the
explicit commands are \texttt{project reset fresh --yes} and
\texttt{project reset state --yes}; ambiguous
\texttt{project reset --yes} is rejected.

\begin{notebox}{How Studio reports outcomes}
A green \(\checkmark\) means an operation or check succeeded. A yellow
warning sign means setup is incomplete or needs attention, but is not
necessarily broken. A red \(\times\) means an operation or validation check
failed. The symbol always carries the meaning, so the output remains clear
without color. Studio adds color only in an interactive terminal; redirected
output and sessions with \texttt{NO\_COLOR} remain color-free. Optional
providers that you are not using are warnings rather than failures.

Every nonempty command also starts a clearly separated output block. For
example:

\begin{lstlisting}[style=shell]
zippergen [reviewed_answer]> current

-- Output: current ----------------------------------------
Current
-------
\end{lstlisting}

The real terminal uses a continuous rule character; the ASCII form above is
print-friendly. The boundary repeats only the command family, never its prompt
text, paths, model values, or secrets. It therefore remains short even for a
long \texttt{create} or \texttt{spec refine} command. Blank input and
\texttt{exit} produce no boundary.
\end{notebox}

\subsubsection*{Step 1a: use commands or ordinary language}

Studio's exact commands remain the durable, scriptable interface, but they do
not have to be memorized. At the Studio prompt, try:

\begin{lstlisting}[style=shell]
What is the current state?
Show me the whole protocol.
Assign the mock model to Writer.
\end{lstlisting}

The first request becomes \texttt{current}; after a workflow is selected, the
second becomes \texttt{show protocol}, and the third becomes
\texttt{models set Writer mock}. Studio prints a
\emph{Natural-language request} table and the exact command plan before any
result, so natural language never hides which operation was chosen. Common
requests use a deterministic interpreter and therefore need no model, network,
or API key.

When a request depends on repository details and no built-in interpretation
matches, Studio may run the already authenticated Codex or Claude Code CLI
once. The CLI receives non-secret project context and read-only repository
access. It may inspect workflow code, \path{specification.md}, and other
project Markdown, but it can return only a structured plan made from documented
Studio commands. Studio validates that plan and performs the operations
itself; model text is never executed as shell code.

The execution rule is intentionally visible. Read-only requests and clear,
reversible configuration changes run directly. Starting or stopping execution,
deployment operations, and destructive changes require confirmation.
\texttt{plan TEXT} always shows an interpretation without executing it;
\texttt{ask TEXT} explicitly interprets and executes prose. A question such as
\texttt{How do I show the protocol?} is also treated as preview-only.

Inspect the language layer at any time:

\begin{lstlisting}[style=shell]
language
language set auto
language set codex       # or: language set claude
language learning on
language history
language learned
language forget L001
\end{lstlisting}

\texttt{auto} prefers Codex and then Claude when a repository-aware fallback
is needed. It reuses the selected CLI's existing login; no separate ZipperGen
provider key is needed. \texttt{language set off} retains deterministic and already learned
interpretations without launching a CLI. After a successful CLI plan, Studio
stores the accepted request, structured commands, and outcome in owner-private
project state. It generalizes recognized values: an interpretation learned for
\texttt{Writer} can later match the same wording for \texttt{Reviewer}.
Nothing is added to the repository and \texttt{project reset state} archives
this history with the other private state. Use \texttt{language learning off}
to stop adding entries or \texttt{language forget ID|all} to remove them.

Raw CLI output is not learned. Secret-looking requests are rejected before
interpretation and are not stored; enter provider credentials through
\texttt{providers set NAME}, which uses Studio's private secret input.

\subsubsection*{Step 1b: choose a terminal editor once}

Studio can open specification and workflow files without making you leave its command
loop. First ask which editor it would use automatically:

\begin{lstlisting}[style=shell]
editor show
\end{lstlisting}

The output separates the remembered preference, effective executable, and its
source. With no preference, Studio checks the ordinary \texttt{\$VISUAL} and
\texttt{\$EDITOR} settings, then looks for \texttt{micro}, \texttt{nano},
\texttt{vim}, and \texttt{vi}, in that order. If \texttt{micro} is installed
and you prefer it, remember that choice for this project:

\begin{lstlisting}[style=shell]
editor set micro
editor show
\end{lstlisting}

Use \texttt{editor set nano} instead if that is the installed editor you want.
The choice is machine-specific Studio state, not a line in the workflow or a
file committed to Git. It survives leaving Studio, reopening the terminal, and
a computer crash. Enter \texttt{editor reset} to return to automatic discovery.

An editor briefly takes over the same terminal. Save the file and exit the
editor to return to the unchanged Studio session. Studio executes the editor
program directly; it does not invoke an LLM, shell script, workflow provider,
or MCP server. A one-time choice can be supplied without changing the saved
preference:

\begin{lstlisting}[style=shell]
edit workflow --editor nano
create --editor micro
\end{lstlisting}

For an editor that needs arguments, quote the whole one-time command, for
example \texttt{--editor "code --wait"}. Studio accepts edited files only
inside the project root, and it does not continue when the editor reports
failure. These checks keep ``edit this file'' scoped to the current project.

\subsubsection*{Step 2: write the canonical specification}

Long requirements do not belong in a single-line terminal question, and the
user should not have to invent or remember a prompt filename. Enter
\texttt{create}. Studio creates the fixed project file
\path{specification.md}, opens it in the editor selected in Step~1b, and
waits. On the first creation it contains a short comment explaining what a
specification should cover and which development mechanics to omit. Write
below that comment or replace it with the complete text below. Studio removes
the guide automatically after real requirements are saved; leaving the guide
untouched does not create an assistant task. Copying into an editor is reliable
even when a terminal's interactive question cannot accept a multiline paste.

\begin{lstlisting}[style=shell]
create
\end{lstlisting}

\begin{lstlisting}[numbers=none]
# Reviewed Answer

Produce a helpful answer to a request, subject to automated review and explicit
human approval.

## Participants and behavior

- Requester initially owns `request` and `max_retries`.
- Writer uses an LLM to draft and revise an answer.
- Reviewer uses a separate LLM to assess every draft.
- After the automated assessment, a human approves or rejects the draft.
- A rejection sends the draft back to Writer while retries remain.
- Never return an unapproved draft. Return an explicit failure after retry
  exhaustion.

## Inputs and operation

- The application accepts a `request` input, defaulting to
  `Explain why the sky is blue in two sentences.`
- It accepts an integer `max_retries` input, defaulting to `2`.
- Use the deterministic mock model by default so the application can be tried
  without a paid model account.
- A deployment may instead use OpenAI or Anthropic. Request the corresponding
  API key only when that provider is selected.
\end{lstlisting}

Save the file and exit the editor. Studio reads the whole UTF-8 document,
preserves its paragraphs and lists, and wraps it in repository-aware
instructions. It prints a compact table instead of repeating the long text:

\begin{lstlisting}[style=shell]
-- Output: create -----------------------------------------
Creation
  Specification  [success] specification.md
  Task           [success] .zippergen/current-task.md
  Next                     assistant codex / assistant claude
  Inspect                  task / task show / task history
\end{lstlisting}

There is no prompt ID, numbered prompt file, or ordering operation. The
canonical document is always \path{specification.md}; it is ordinary project
source and should be reviewed and committed with the Python workflow. The
generated, non-secret handoff is always
\path{.zippergen/current-task.md}; Studio also keeps timestamped private task
history outside Git. The task asks the assistant to identify participants,
ownership, messages, action kinds, decisions, loops, deployment needs, safety
assumptions, and acceptance examples before editing. It also requires tests,
validation, semantic views, and all exact local projections. No workflow file
has been created yet; Step~3 starts the coding assistant on this task.

Notice that the specification does not choose a Python filename or symbol,
request tests and validation, or repeat the prohibition on deployment. Those
are development mechanics supplied automatically by Studio's generated task.
The specification contains only durable application intent. After generation,
Studio discovers the resulting workflow so the user can select it with
\texttt{use} without knowing its source path in advance.

The everyday specification commands are intentionally uniform:

\begin{lstlisting}[style=shell]
spec show       # read the accepted requirements
spec edit       # reopen the same specification.md
spec path       # print its absolute path only when another tool needs it
task            # inspect the generated assistant-task summary
task show       # inspect the complete assistant instruction
task history    # list timestamped private handoffs
\end{lstlisting}

An empty specification is rejected. Never put an API key or other secret in
it. For a genuinely short experiment, \texttt{create DESCRIPTION} writes the
same canonical file without opening an editor. Advanced users may import an
existing UTF-8 document once with \texttt{create --file PATH}; Studio copies
its content into \path{specification.md} and does not retain that source
filename as project configuration.

Studio fingerprints the canonical specification when it prepares a task.
While the task is still \texttt{ready for assistant}, Studio compares that
fingerprint before \texttt{task}, \texttt{task show}, \texttt{task path},
\texttt{current}, or \texttt{assistant}. If an input document changed, Studio
creates one refreshed private task and records which older request it replaces.
After an assistant has run, its expected edits are output awaiting review, not
new task input: Studio preserves the executed request instead of silently
creating another ``ready'' task.

Projects made with an older Studio may still contain
\path{prompts/index.toml} and numbered prompt files. On the first
\texttt{spec show}, \texttt{spec edit}, or \texttt{spec refine}, Studio safely
combines their active contents into \path{specification.md} and keeps every old
file as history. New work then uses only the canonical specification model.

\begin{notebox}{Why Studio creates a brief instead of hidden code}
Natural language is flexible; executable coordination must be precise. The
brief supplies repository rules and acceptance checks. The resulting ordinary
Python file is inspected and versioned. ZipperGen never treats opaque model
output as an automatically deployed workflow.
\end{notebox}

\subsubsection*{Step 3: choose an assistant and execute the task}

At the same Studio prompt, enter:

\begin{lstlisting}[style=shell]
assistant codex
# Or, when Claude Code is installed:
assistant claude
\end{lstlisting}

Choose one of the two commands; do not run both for the same task. Plain
\texttt{assistant} remains an alias for \texttt{assistant codex}. Studio runs
the selected local coding assistant once in the project root and tells it to
read \path{.zippergen/current-task.md}. Just before launch, Studio applies
the freshness check described above, so the file contains the current
canonical specification. There is no task path or prompt file to copy. The
assistant autonomously inspects and edits the application project within its
configured permission boundary, runs checks, writes a small structured
verification handoff, exits, and returns control to the same ZipperGen Studio
session. The handoff distinguishes the assistant process returning from the
requested checks actually passing. Use
\texttt{assistant codex --interactive} only when a conversation or interactive
approval is genuinely needed.

The generated task also explains the tutorial's two-project layout. The
application lives at the tutorial root, while \path{zippergen/pyproject.toml}
defines the environment for the nested framework checkout. It therefore tells
the assistant to invoke the framework from the application root with
\texttt{uv run --offline --project zippergen zippergen ...} and to run
application tests by the explicit \path{tests} path. A bare
\texttt{uv run pytest} at the tutorial root is deliberately forbidden: it can
collect \path{zippergen/tests} while
using an environment that did not install the framework dependencies. The
framework's own suite is separate and is needed only when framework source was
actually changed. Pytest is a declared development dependency installed by the
initial \texttt{uv sync}; the assistant must not add
\texttt{--with pytest}, install packages, or request PyPI access. If the local
environment has not been synchronized after an update, run
\texttt{uv sync --project zippergen} once in the ordinary terminal and then
rerun the assistant deliberately. Studio checks this prerequisite offline
before starting Codex or Claude, so a missing pytest installation does not
consume an assistant run.

For one-shot Codex runs, Studio uses Codex's structured JSON output mode and
shows the final assistant report rather than its raw execution trace. Known
internal model-cache diagnostics are suppressed; unexpected diagnostics are
condensed into warnings. Interactive Codex sessions retain their ordinary
terminal output.

There is no assistant queue or scheduled background job. The command runs
synchronously: before launch, \texttt{task} says
\texttt{ready for assistant} and
\texttt{Execution: not started; nothing is scheduled}. A normal return
changes the same request to \texttt{awaiting human review} and replaces the
assistant suggestion with the concrete review path. Separately,
\texttt{Verification} is \texttt{passed}, \texttt{failed}, or
\texttt{incomplete}, with a concise note and the number of checks the
assistant reported. The \texttt{task} output lists each reported command and
its outcome underneath the summary. A missing, malformed, or internally
inconsistent handoff is never interpreted as success. A failed or interrupted
assistant process remains visible and retryable. If Studio or the computer
stops while the assistant is running, the orphaned state is recovered as
\texttt{assistant interrupted} on the next inspection. Studio refuses to start an
\texttt{awaiting human review} task accidentally; use
\texttt{assistant codex --rerun} or
\texttt{assistant claude --rerun} only when a second pass is intentional.
Studio refreshes an older unexecuted task automatically before launch so that
it also receives this verification contract.

\begin{lstlisting}[style=shell]
task
Current task
  Status     [warning] awaiting human review
  Assistant            Codex -- returned ...
  Execution            assistant session ended; nothing is scheduled
  Verification [success] passed -- 5 checks
  Verification note    Validation, projections, diff, and focused tests passed.
  Context    [success] assistant result is preserved for review
  Next                  use / current / validate / show / task close
\end{lstlisting}

The verification result is an explicit assistant report, not an independent
proof and not permission to deploy. If it is \texttt{failed} or
\texttt{incomplete}, Studio shows that state prominently and proposes
inspection, validation, and an intentional assistant rerun instead of implying
that the change is ready. Immediately below the review table it prints only
failed or unexecuted checks; the complete reported check list remains
available through \texttt{task}. Human review of the visible code and semantic
views remains mandatory.

For this initial creation task, use \texttt{task close} after the generated
workflow has passed the review below. The command removes only the stable
\path{.zippergen/current-task.md} mirror; timestamped private history remains.
Refinement tasks close differently: their explicit acceptance operation is
\texttt{spec reconcile}, while \texttt{spec discard} records rejection.

If Studio says that \texttt{codex} is missing, the prepared task remains safe.
Open a second operating-system terminal. If npm is already installed, install
and authenticate Codex once:

\begin{lstlisting}[style=shell]
npm install --global @openai/codex
codex login
cd "$HOME/Projects/zippergen-tutorial"
codex
\end{lstlisting}

Then return to Studio and enter \texttt{assistant codex} again.
Section~\ref{sec:codex} explains setup for both supported assistants and the
no-assistant fallback in more detail. The task points
the assistant to
\path{zippergen/AGENTS.md} and the nested framework's workflow skill when the
skill is not installed at the parent project root. The long workflow
specification remains in \path{specification.md} rather than being pasted into
the assistant's terminal.

\begin{notebox}{What \texttt{assistant} is---and is not}
\texttt{assistant codex} and \texttt{assistant claude} are small launchers for
the locally installed Codex CLI and Claude Code respectively. They do not send
the task through ZipperGen's Writer or Reviewer model providers and need no
ZipperGen API key. Each tool uses its own authentication, model settings,
approval policy, and tools. MCP is not required. Independently configured MCP
servers remain available to the chosen assistant, but Studio neither creates
nor manages them. Another repository-aware assistant can instead read
\texttt{task show} or \texttt{task path}.
\end{notebox}

Wait for the assistant to report generated source, focused tests, validation,
communication-only and full views, and projections for Requester, Writer, and
Reviewer. Then inspect the checkout in the ordinary terminal:

\begin{lstlisting}[style=shell]
git status --short
git diff
\end{lstlisting}

Do not commit merely because code was generated. First complete the checks
below. If generation is unavailable or fails, continue with Step~4 and select
the shipped \texttt{tutorial\_review} workflow instead. The selector displays
its internal source reference; the user does not have to type it.

\subsubsection*{Step 4: select and validate the generated workflow}

Return to the Studio terminal and enter:

\begin{lstlisting}[style=shell]
use
current
validate
\end{lstlisting}

Studio displays the discovered workflows. Select the newly generated
reviewed-answer workflow by its number; there is no path or Python symbol to
type. The selected entry may display an internal reference of the form
\texttt{PATH.py:NAME}: the suffix names the Python workflow object, not a
deployment. The three commands have deliberately different jobs:

\begin{description}
  \item[\texttt{use}] discovers top-level workflow objects, loads the chosen
        object, and remembers its \texttt{PATH.py:NAME} reference as the current
        workflow for later Studio commands. It neither runs nor deploys it.
  \item[\texttt{current}] rebuilds the project dashboard from the current files
        and machine-specific Studio state. Once a workflow is selected, this
        includes a fresh validation summary, all participants, an explicit
        \texttt{LLM-active participants} subset derived from actual
        \texttt{@llm} action sites, model routing, provider configuration, the
        managed run, and deployment. Thus the green
        \texttt{Validation: valid} mark immediately after \texttt{use} is
        expected: \texttt{current} has checked the selected source; it is not
        a success mark cached by \texttt{use}.
  \item[\texttt{validate}] runs the validator explicitly and prints every
        individual check, including the exact local projection checks. It does
        not modify, execute, or deploy the workflow.
\end{description}

\texttt{current} should show no run, no deployment, and default LLM
\texttt{mock}. In the detailed \texttt{validate} output, every successful check
begins with a green check mark. A red cross identifies a failed check and is
followed by its explanation.

\subsubsection*{Step 5: inspect increasing degrees of detail}

Enter \texttt{show} and choose each of the first five views once:

\begin{lstlisting}[style=shell]
-- Output: show -------------------------------------------
1. Overview
2. Protocol
3. Communications only
4. Actions and prompts
5. Complete workflow
6. One participant
7. Selected participants
\end{lstlisting}

Then use the direct forms:

\begin{lstlisting}[style=shell]
show communications
show actions
show full
show agent Writer
show agent Reviewer
show agents Writer Reviewer
\end{lstlisting}

All results are code. \texttt{show communications} hides local actions and
retains messages and control. \texttt{show actions} reveals action calls and
prompts. \texttt{show full} adds implementation and deployment requirements.
\texttt{show agent Reviewer} is the exact projected local program.
\texttt{show agents Writer Reviewer} is a focused global view with hidden
peers marked as boundaries; it is not a projection.

Before running, confirm from these views who owns both inputs, who owns the
retry decision, whether every revision returns to human review, whether an
unapproved draft can reach Requester, and whether both model prompts are
visible. Do not continue until the code answers all five clearly.

\subsubsection*{Step 6: configure models without shell exports}

Enter:

\begin{lstlisting}[style=shell]
models reset all
models show
models check
\end{lstlisting}

Studio lists only lifelines containing \texttt{@llm} action sites, together
with their action names. Requester is absent because message passing and pure
validation do not require a model. Expected shape:

\begin{lstlisting}[style=shell]
Models for reviewed_answer
  Default: mock
  Writer: mock (inherits default; actions: draft_reply, revise_reply)
  Reviewer: mock (inherits default; actions: assess_draft)
\end{lstlisting}

Generated action names may differ, but Writer and Reviewer should appear. The
non-secret profile is stored per workflow and survives a reboot.

\subsubsection*{Step 7: make a durable mock run in one terminal}

Enter \texttt{run}. Studio validates and asks for each workflow input:

\begin{lstlisting}[style=shell]
zippergen [reviewed_answer]> run
Workflow reviewed_answer: valid
Request to answer (str) ["Explain why the sky is blue in two sentences."]:
Maximum revisions after the initial draft is rejected (int) [2]: 3
Run reviewed_answer-<timestamp>

Draft:
[draft_reply:draft]

Automated reviewer notes:
[assess_draft:concerns]

Approve this draft? [y/n]: n
\end{lstlisting}

Press Enter for the displayed request, enter \texttt{3} for the retry budget,
\texttt{n} for the first review, and \texttt{y} for the second. The mock result
is a recognizable placeholder. The important behavior is a separate Writer
draft, Reviewer assessment, visible human authority, and another revision and
assessment after rejection.

Enter \texttt{current} and \texttt{runs}. Studio shows a unique managed SQLite
store even though no filename was typed. Every new \texttt{run} receives a new
store and never silently reuses an old result.

\subsubsection*{Step 8: verify crash and resume behavior}

Start another run and press Control-C at the human prompt. Studio records the
managed run as interrupted and prints a recovery hint instead of a traceback.
The expected transition is:

\begin{lstlisting}[style=shell]
Approve this draft for the requester? [y/n]: ^C
Command interrupted. Use 'current' to inspect context; use 'resume' for an incomplete managed run.
zippergen [reviewed_answer]>
\end{lstlisting}

Only Studio's main thread reads the terminal. After the warning, no abandoned
workflow thread should consume a Studio command or print another approval
prompt. You may enter \texttt{current} and \texttt{resume} immediately;
\texttt{current} reports the run as interrupted, and \texttt{resume} presents
the same still-pending approval once. To test a full terminal or computer
restart, leave Studio, reopen a terminal later, and enter:

\begin{lstlisting}[style=shell]
cd "$HOME/Projects/zippergen-tutorial"
zippergen
current
resume
\end{lstlisting}

\texttt{resume} reuses recorded inputs, model routing, and SQLite state; the
pending task is presented again. It also checks the workflow's semantic
fingerprint. If source meaning changed, Studio preserves the store and refuses
an unsafe resume.

No export must be repeated. The workflow, model routing, run record, and
development secrets live below \path{$HOME/.zippergen/workspaces} by default.
A shell export disappears with its shell; Studio's workspace record does not.

The initial creation task has now been generated, inspected, validated, run,
and recovery-tested. Close that development task explicitly:

\begin{lstlisting}[style=shell]
task
task close
\end{lstlisting}

\texttt{task} should say \texttt{awaiting human review} and show
\texttt{task close} in its \texttt{Next} row. Confirm the close when Studio
asks. This does not remove the specification, workflow, tests, run, or private
task history; it only records that this creation task is no longer open.
\texttt{current} will then show no current development task. The optional
model-routing step below and later deployment do not require an open assistant
task.

\subsubsection*{Optional Step 9: route two lifelines to different real models}

Only do this after the mock path succeeds and only if you have provider
accounts, valid model access, and permission to incur API charges. Otherwise
skip directly to Step~10; the workflow remains fully usable with mock models.

If you choose this optional step, prepare each provider account before changing
the Studio profile:

\begin{enumerate}
  \item Sign in to the provider's API console, not merely its consumer chat
        application. OpenAI keys are managed at
        \url{https://platform.openai.com/api-keys}; Anthropic API access and
        keys are managed through \url{https://console.anthropic.com}.
  \item Select or create a development project/workspace and confirm that the
        account has API billing or credits. API usage may be billed separately
        from a consumer chat subscription; do not assume the latter funds it.
  \item Set a small development budget, usage alert, or credit balance where
        the provider supports it. Confirm that the intended model is enabled
        for that project/workspace.
  \item Create a separate development key. A provider may show the complete
        secret only once. Keep it private and available only long enough to
        paste it into Studio's hidden prompt below. Do not type it into a shell
        command to test it.
\end{enumerate}

See OpenAI's current
\href{https://help.openai.com/en/articles/4936850}{API-key location} and
\href{https://help.openai.com/en/articles/5112595-best-practices-for-api-key-safety}{API-key safety guidance}.
If a key is lost, pasted into the wrong place, or appears in Git output, revoke
it in the provider console and create a replacement before continuing.

\begin{minipage}{\linewidth}
\begin{lstlisting}[style=shell]
providers
providers set openai
providers set anthropic
models default mock
models set Writer openai:gpt-4o-mini
models set Reviewer claude:claude-sonnet-4-6
models show
models check
\end{lstlisting}
\end{minipage}

Each \texttt{providers set} command asks for its API key without echo and saves
it in owner-only Studio storage. Plain \texttt{providers} makes no remote
request and never prints a secret. An API provider is \texttt{configured} when
its key is present, and its line explicitly says \texttt{not tested here}. A
local provider line reports the last saved check result and timestamp. For a
local OpenAI-compatible or Ollama endpoint use
\texttt{providers set local URL}, followed by a model such as
\texttt{ollama:qwen2.5:7b}. Studio immediately requests the endpoint's
OpenAI-compatible \texttt{/models} route with a three-second timeout. It saves
the URL only when the response is valid, then reports the check time and number
of available models. Enter \texttt{providers check local} after reconnecting
an SSH tunnel or restarting the model server. A failed recheck changes the
display to \texttt{unreachable} and explains that the server or tunnel should
be checked. The model identifiers above are examples; use models enabled for
your provider accounts when necessary.

Studio verifies model routing at the point where it is entered. After a
provider has been configured, \texttt{models default SPEC} and
\texttt{models set LIFELINE SPEC} query that provider's model endpoint using
the private key. A green check means that the exact model identifier or alias
is available to that key. A name that the provider rejects is not saved, so a
typing error cannot silently survive until the first workflow action. Local
identifiers are compared with the endpoint's live \texttt{/models} response.
If a configured provider is temporarily unreachable, Studio keeps the
deliberately chosen routing but prints a yellow \emph{saved but unchecked}
warning; the provider will still enforce the selection when the workflow runs.
If no key has been configured yet, Studio likewise saves the non-secret route
with a warning and tells you which \texttt{providers set} command is needed.

\texttt{models check} is the read-only recheck command. With no argument, or
with \texttt{all}, it checks the default plus every LLM-active participant.
\texttt{models check Reviewer} checks only Reviewer's effective assignment;
\texttt{models check default} checks only the inherited default. Studio queries
an identical effective route only once, labels every assignment covered by the
result, and finishes with \texttt{Routing: unchanged}. A green result confirms
that the provider accepted the configured key and exact model identifier at
check time. A yellow result means connectivity or configuration prevented a
conclusive check; it is not proof that a generation call will succeed later.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[node distance=13mm and 18mm]
  \node[zgplain, text width=28mm] (default) {Default\\\texttt{mock}};
  \node[zgstage, text width=34mm, below left=of default] (writer)
    {Writer actions\\\texttt{openai:...}};
  \node[zgstage, text width=34mm, below right=of default] (reviewer)
    {Reviewer actions\\\texttt{claude:...}};
  \node[zgplain, text width=31mm, below=of writer] (okey)
    {\texttt{OPENAI\_API\_KEY}\\private storage};
  \node[zgplain, text width=34mm, below=of reviewer] (akey)
    {\texttt{ANTHROPIC\_API\_KEY}\\private storage};
  \draw[zgflow] (default) -- node[zglabel, left]{override} (writer);
  \draw[zgflow] (default) -- node[zglabel, right]{override} (reviewer);
  \draw[zgdashed] (okey) -- (writer);
  \draw[zgdashed] (akey) -- (reviewer);
\end{tikzpicture}
\caption{One workflow routes LLM-active lifelines independently. Model names
are configuration; provider keys are separate secrets.}
\label{fig:model-routing}
\end{figure}

The provider commands display lines such as:

\begin{lstlisting}[style=shell]
OPENAI_API_KEY:
ANTHROPIC_API_KEY:
Configured anthropic: configured; ANTHROPIC_API_KEY is in private Studio storage; not tested here
\end{lstlisting}

The configured line begins with a green check mark; unconfigured optional
providers begin with a yellow warning sign. Paste each key only at the hidden
Studio prompt---never into source, a shell command, an assistant prompt, or a
Git-tracked \path{.env}. Development secrets are mode-0600 and absent from
workspace, run, and request JSON. Existing environment values are used but not
copied. Later development runs temporarily inject the selected configured
providers and restore the process environment afterward. Use
\texttt{providers reset NAME} to remove a saved configuration and private key.

Use \texttt{models reset Writer} to restore inheritance for one role, or
\texttt{models reset all} for the declared default. The scriptable equivalent
is repeatable \texttt{--llm-for}:

\begin{lstlisting}[style=shell]
zippergen dev workflows/reviewed_answer.py:reviewed_answer \
  --llm mock \
  --llm-for Writer=openai:gpt-4o-mini \
  --llm-for Reviewer=claude:claude-sonnet-4-6
\end{lstlisting}

\begin{notebox}{Two different meanings of ``change the model''}
For a routing-only change, use \texttt{models default SPEC} or
\texttt{models set LIFELINE SPEC}. Studio remembers that choice for runs and
deployment; source code does not change, so no refinement or coding assistant
is needed. If the model choice must become versioned design intent, change
declared workflow/deployment defaults, or accompany prompt, action, or test
changes, record it as a refinement and run an assistant. For example:

\begin{lstlisting}[style=shell]
refine Use openai:gpt-4o-mini for Writer and preserve all protocol behavior.
assistant claude
\end{lstlisting}

The second form appends to the one pending refinement and asks the assistant to
update the canonical specification and visible project files together. It is
intentionally heavier than routing.
\end{notebox}

\subsubsection*{Step 10: refine through one pending change}

If Step~9 used paid models, return to deterministic mock execution before
changing behavior unless another paid run is intentional:

\begin{lstlisting}[style=shell]
models reset all
\end{lstlisting}

This changes only routing; it does not delete saved development secrets. Enter:

\begin{lstlisting}[style=shell]
spec refine
\end{lstlisting}

Studio creates the automatically named, ignored file
\path{.zippergen/pending-refinement.md} and opens it. Paste this change, save,
and exit the editor:

\begin{lstlisting}[numbers=none]
Add a FactChecker participant. Reviewer must request a factual check before
the human decision. Preserve the bounded retry and approval behavior. Add
focused tests and do not deploy.
\end{lstlisting}

There is only one pending refinement. Entering \texttt{spec refine} again
reopens that same file; it never asks for a name and never creates P002, P003,
or another permanent prompt fragment. The short \texttt{refine} command is a
first-class alias. Supplying short text, for example \texttt{refine Also keep
the mock backend as the default}, appends it to the existing pending document.
\texttt{spec pending} displays the complete change without opening the editor.
Appending here never appends a chronological note to
\path{specification.md}.

This step is useful because creation and later modification must follow the
same safe cycle; it is not required merely to deploy the initially generated
workflow. Studio saves one semantic pre-edit baseline and creates a task
containing the accepted \path{specification.md}, the pending refinement, the
selected workflow, and that baseline. Enter \texttt{task} if you want to
inspect it, then \texttt{assistant} as in Step~3. The assistant is explicitly
required to integrate the change coherently into \path{specification.md},
rather than turning that document into a chronological journal; it updates
code and tests, validates changed projections, and runs
\texttt{zippergen diff}. It must leave the pending file in place for human
review.

This is the complete repeatable change loop; use it for every later behavioral
or source-level modification:

\begin{lstlisting}[style=shell]
spec refine            # create or reopen the one pending change
spec pending           # optional: inspect it without editing
task                 # optional: inspect the generated handoff
assistant codex      # or: assistant claude
current
validate
show communications
show agent Writer    # choose every participant changed by the refinement
run
spec show             # verify the assistant integrated the requirement
spec reconcile        # confirm, archive, and clear the pending change
\end{lstlisting}

The refinement must come before the assistant call. Before the first assistant
launch, editing either input document refreshes the prepared task
automatically. After the assistant returns, \texttt{task} reports
\texttt{awaiting human review}; it does not reinterpret the expected edits as
a new task. Its \texttt{Next} row names the review commands above. Nothing is
scheduled, and a second assistant launch requires an explicit
\texttt{--rerun}.

\texttt{spec reconcile} is an acceptance and cleanup operation, not a merge
engine. The coherent fusion happens either manually with \texttt{spec edit} or
through \texttt{assistant codex}/\texttt{assistant claude}. Simple copying is
possible in an editor but is usually the wrong result: the canonical document
should describe the current application cleanly rather than accumulate update
notes.
\texttt{spec reconcile} refuses to clear the pending document when the
canonical specification has not changed since refinement began; this catches
an assistant that changed code but forgot the durable requirements. It asks
for human confirmation, accepts the already-integrated canonical document,
moves the pending text into recoverable private history, and closes the task.
Use \texttt{spec discard} for an unwanted change and \texttt{spec history} to
list reconciled and discarded refinements. Accepted specification history
remains visible in Git.

\subsubsection*{Step 11: perform the pre-deployment review}

Enter \texttt{show full}. In the ordinary terminal run:

\begin{lstlisting}[style=shell]
git status --short
git diff --check
git diff
\end{lstlisting}

Confirm that the full view declares both workflow inputs, mock default, both
conditional provider secrets, and the generated file as a bundled file. Check
that tests pass and no key appears in the diff. Commit the reviewed manifest,
canonical specification, workflow source, and tests; private Studio state and
secrets remain outside the checkout.

Run the project tests and the type checker. The assistant chose the individual
test filename, so the tutorial does not require it:

\begin{lstlisting}[style=shell]
uv run --offline --project zippergen python -m pytest -q tests
uvx pyright
\end{lstlisting}

The test command should end with a passing count. \texttt{uvx pyright} may
download the type-checker on first use and should end with zero errors; skip it
temporarily only when network policy blocks that download, and record the
unverified check.

If both succeed, stage the visible application contract, generated workflow
directory, and tests, inspect the staged diff, and commit:

\begin{lstlisting}[style=shell]
git add zippergen.toml specification.md workflows tests
git diff --cached
git commit -m "Add reviewed answer workflow"
\end{lstlisting}

The commit is a local recovery point. Pushing is a separate choice and is not
required for local deployment.

\subsubsection*{Step 12: deploy and operate the reviewed workflow}

First prepare the named deployment without starting a background service. This
command snapshots the declared source, creates a deployment-specific Python
environment, writes service definitions, and runs readiness checks. It changes
files below ZipperGen's private home but deliberately defers service-manager
state. On the tutorial machine, enter in Studio:

\begin{lstlisting}[style=shell]
deploy reviewed-answer-tutorial --no-start
\end{lstlisting}

Studio carries the current default and per-lifeline routes into the deployment
profile. If the current model profile selects a real provider and its key was
configured earlier, Studio displays a \emph{Deployment credentials} summary
containing the key's name but never its value:

\begin{lstlisting}[style=shell]
Deployment credentials
  Available   [success] OPENAI_API_KEY in private Studio storage
  Deployment             reviewed-answer-tutorial
  Storage                separate private deployment secret file

Reuse the configured credential for deployment reviewed-answer-tutorial? [Y/n]:
\end{lstlisting}

Press Enter to reuse it. Studio copies the hidden value directly into the
deployment's separate mode-0600 file, so the guided deployer does not request
the key again. Enter \texttt{n} only when this deployment should use a different
credential; the deployer then asks for that value without echo. A later
\texttt{deploy reviewed-answer-tutorial} retains the deployment's existing key
and prints that it is keeping the saved value, without presenting another
secret prompt. With the recommended \texttt{models reset all} above, all routes
are mock and no provider key is needed.

Read every prompt. The request and retry budget are runtime inputs, not another
description of the workflow. A successful ending prints
\texttt{Deployment: reviewed-answer-tutorial} and leaves the service stopped.
Studio remembers the deployment name. Check the prepared artifacts before the
first real service-state change:

\begin{lstlisting}[style=shell]
doctor
status
\end{lstlisting}

Doctor should contain no failures; before first execution, a missing store/log
and an uninstalled service are expected warnings. If the checks are acceptable
and service startup is authorized, enter these one at a time:

\begin{lstlisting}[style=shell]
start
status
logs
restart
status
stop
\end{lstlisting}

\texttt{start} is the first command here that changes service-manager state.
The process may need a few seconds to reach its first action; repeat
\texttt{status} once if necessary. It should report \texttt{waiting} once
durable human approval is pending. \texttt{logs} shows the service output and
returns;
\texttt{restart} deliberately proves that the same
pending task survives a process restart. Then \texttt{stop} prevents this
finite tutorial workflow from being supervised indefinitely. Later chapters
show foreground execution, task tokens, and notification adapters for actually
answering the out-of-process approval. Stopping does not delete the profile,
bundle, log, secrets, or store.

If Studio reports that no supported user service manager is available, stop at
the error and use the later \texttt{deploy --no-start} plus
\texttt{run-deployment} foreground path. Do not repeatedly retry a failing
service start without reading \texttt{doctor} and \texttt{logs}.

\begin{warningbox}{Deployment is a real operational action}
First complete mock execution and inspect \texttt{show full}. A real model call
may cost money and an \texttt{@effect} may change an external system. Use fake
services until effects and retry safety have been reviewed.
\end{warningbox}

\subsection{Alternative track: use the shipped running example}

Use this track if you skipped Codex or want to rehearse Studio against a known,
versioned result before generating anything. It intentionally repeats the
inspection, mock run, resume, model, and deployment controls using
\path{zippergen/examples/tutorial_review.py}. If you completed the generated-workflow
track above and those controls are already clear, skip to
Section~\ref{sec:reference-chapters}.

From the project root, open Studio with the tutorial workflow already
selected:

\begin{lstlisting}[style=shell]
zippergen studio zippergen/examples/tutorial_review.py:tutorial_review
\end{lstlisting}

The beginning of the session looks like this; the absolute project path will
of course be different on your machine:

\begin{lstlisting}[style=shell]
ZipperGen Studio
Project: /path/to/zippergen-tutorial
Workflow: zippergen/examples/tutorial_review.py:tutorial_review
Type 'help' for commands; 'show' opens the inspection menu.
zippergen [tutorial_review]>
\end{lstlisting}

The text in square brackets is the current workflow object name, not a
deployment name. Enter \texttt{help} at any time. The important first commands
are:

\begin{lstlisting}[style=shell]
project show   # visible manifest and automatic specification path
current        # specification, workflow, models, run, deployment
editor show    # effective terminal editor and preference source
editor set micro # remember an installed editor for this project
spec show      # canonical accepted requirements
spec edit      # edit the same automatically named specification.md
spec refine    # create/reopen the one pending refinement
spec pending   # inspect that change without opening the editor
spec reconcile # accept an integrated change and clear pending state
task            # lifecycle: prepared, running, review, failed/interrupted
task close      # close a reviewed creation task; history remains
use            # discover workflows and select one from a numbered list
show           # choose a semantic code view from a numbered list
validate       # validate the selected workflow and every projection
models         # configure default and per-lifeline LLMs
models check   # recheck effective model availability without changing routes
providers      # configuration/last local check; no remote API request
providers check local # recheck a configured local model server or SSH tunnel
run            # fresh guided, durable development run
resume         # continue the current incomplete managed run
runs           # list this project's managed development runs
project reset  # choose fresh design, state-only reset, or cancel
create          # create/reopen specification.md and prepare a brief
refine          # short alias for spec refine
assistant codex # execute the current task; assistant claude is also supported
assistant codex --rerun # deliberately repeat a task awaiting review
deploy         # explicitly enter guided named deployment
exit           # leave Studio
\end{lstlisting}

These are Studio commands entered after the
\texttt{zippergen [tutorial\_review]>} prompt. Do not prefix each one with
\texttt{zippergen} while already inside Studio.

\subsection{Inspect the workflow as code}

Enter \texttt{show}. Studio displays:

\begin{lstlisting}[style=shell]
1. Overview
2. Protocol
3. Communications only
4. Actions and prompts
5. Complete workflow
6. One participant
7. Selected participants
\end{lstlisting}

Choose a number. Every result is code derived from the semantic workflow; the
menu is only a convenient selector. It does not introduce a separate diagram
format. You can also name a view directly:

\begin{lstlisting}[style=shell]
show overview
show protocol
show communications
show actions
show full
show agent Reviewer
show agents Writer Reviewer
\end{lstlisting}

\texttt{show agent Reviewer} is the exact program produced by formal
projection for Reviewer. \texttt{show agents Writer Reviewer} is instead a
focused global view that retains those participants and marks hidden peers as
explicit boundaries. The distinction is explained fully later in the manual.

\subsection{Run durably in the same terminal}

Enter \texttt{run}. Studio validates first and asks for each declared workflow
input, including its type and default:

\begin{lstlisting}[style=shell]
zippergen [tutorial_review]> run
Workflow tutorial_review: valid
Request to answer (str) ["Explain why the sky is blue in two sentences."]:
Maximum revisions after the initial draft is rejected (int) [2]: 3
Run tutorial_review-<timestamp>

Draft:
[draft_reply:draft]

Automated reviewer notes:
[assess_draft:concerns]

Approve this draft? [y/n]: n

Draft:
[revise_reply:draft]

Automated reviewer notes:
[assess_draft:concerns]

Approve this draft? [y/n]: y
Result: [revise_reply:draft]
Next: show another view, start a new run, or prepare deployment.
\end{lstlisting}

Press Enter at the first input prompt to accept its displayed default. The mock
LLM deliberately returns recognizable placeholders. Rejecting the first draft
creates and presents the next durable human task in the same terminal; no
second terminal and no \texttt{tasks} or \texttt{approve} command is required.

The run still uses SQLite. Studio creates a new, unique store and a small JSON
run record below this project's managed workspace. Enter \texttt{current} to
see the exact managed path, or \texttt{runs} to see earlier runs. Starting a
new \texttt{run} never silently reuses an old result.

\subsection{Return after a closed terminal or computer crash}

The current workflow and run are remembered outside the Git checkout. Return
to the same project root and reopen Studio:

\begin{lstlisting}[style=shell]
cd "$HOME/Projects/zippergen-tutorial"
zippergen
\end{lstlisting}

Then inspect and continue:

\begin{lstlisting}[style=shell]
current
resume
\end{lstlisting}

\texttt{resume} reuses the recorded inputs and SQLite store. If a human task
was pending, it is presented again in the terminal. Studio also compares the
current workflow's semantic fingerprint with the one recorded when the run
started. If the workflow meaning changed meanwhile, it preserves the old
store and asks you either to restore matching source or start a new run.

No export needs to be repeated after the crash. By default, Studio uses
\path{$HOME/.zippergen/workspaces}. The later reference exercise deliberately
overrides \texttt{ZIPPERGEN\_HOME} so that every low-level artifact is visible
in one tutorial directory; that override is educational and optional, not a
normal Studio prerequisite.

\subsection{Start from a prompt, or refine an existing workflow}

For a multiline description, let Studio open the canonical project document:

\begin{lstlisting}[style=shell]
create
\end{lstlisting}

For a short idea, provide the description on the same line:

\begin{lstlisting}[style=shell]
create Draft an answer and require human approval.
\end{lstlisting}

Studio saves a structured coding-assistant task. It asks for visible Python
source, focused mock/fake tests, validation, communication and full views,
exact participant projections, and no deployment. The stable generated mirror
is \path{.zippergen/current-task.md}, which is ignored by Git; timestamped
archives remain outside the checkout.

\begin{notebox}{Current assistant boundary}
Studio's \texttt{assistant codex} and \texttt{assistant claude} commands launch
the selected local coding tool on the current task. Neither is a ZipperGen
workflow model provider, and neither automates deployment. The assistant edits
ordinary files in the checkout; ZipperGen then loads, validates, projects,
renders, and runs the result. Other coding assistants can read the same task
through \texttt{task show} or \texttt{task path}. After source creation, enter
\texttt{use} to discover and select the new workflow. The task is not a queued
job: Studio records the synchronous assistant return as
\texttt{awaiting human review}, and \texttt{task close} closes the reviewed
creation handoff without deleting its history.
\end{notebox}

For an existing selected workflow, open or reopen the one pending change:

\begin{lstlisting}[style=shell]
spec refine
\end{lstlisting}

Studio owns both filenames, saves a semantic pre-edit baseline, and includes
the canonical specification plus pending refinement in the assistant brief.
The assistant is required to update \path{specification.md}, preserve behavior
not mentioned in the request, validate the result, inspect changed projections,
and run \texttt{zippergen diff} against that baseline. After review,
\texttt{spec reconcile} accepts the already-integrated canonical
specification, archives the pending transaction, and closes the task. It does
not itself concatenate or rewrite the two documents.

\subsection{Use a real LLM without repeating an API-key export}

Remain on the mock backend until the complete local path works. If you did not
read Optional Step~9 above, first follow its account, billing, key creation, and
revocation precautions. When ready, configure one model for Writer and another
for Reviewer:

\begin{lstlisting}[style=shell]
models set Writer openai:gpt-4o-mini
models set Reviewer claude:claude-sonnet-4-6
models show
models check
run
\end{lstlisting}

Studio remembers these routes per workflow. Because the tutorial declares
conditional required secrets for both selected providers, it asks for both
without echoing either value:

\begin{lstlisting}[style=shell]
OpenAI API key:
Saved OPENAI_API_KEY in private development secret storage.
Anthropic API key:
Saved ANTHROPIC_API_KEY in private development secret storage.
\end{lstlisting}

The secrets are stored once below
\path{$HOME/.zippergen/workspaces/<project>/}, in a file named
\path{development.secrets.json} with owner-only mode-0600 permissions. Later
runs and post-crash resumes reuse it.
The values are not copied into \path{workspace.json}, a run record, an assistant
brief, source code, or Git. Values already supplied by the process environment
are used temporarily but are not copied into the private file.

This is lightweight local secret storage, not an operating-system keychain or
enterprise secret manager. Deployment secrets are scoped separately. When a
selected real provider has a matching key in private Studio storage, the first
deployment explicitly offers to reuse it. Pressing Enter accepts the default;
Studio transfers the hidden value directly into that deployment's own private
mode-0600 secrets file. It never displays the value or puts it in project
state. Answering \texttt{n} preserves the option to enter a distinct deployment
credential. A later deployment with the same name keeps its already-saved
deployment key without asking for it again.

\subsection{Move explicitly from development to deployment}

Deployment may install packages, run declared setup such as OAuth, create a
source bundle and managed environment, and start a macOS or Linux user service.
It is therefore always a separate, explicit command. With the tutorial still
selected, return to mock unless a real deployment is intentional, then choose
a separate name:

\begin{lstlisting}[style=shell]
models reset all
deploy tutorial-review-studio --no-start
\end{lstlisting}

The same guided deployer described later carries the current default and
per-lifeline model routes, collects declared settings and deployment-scoped
secrets, validates readiness, and prints the resulting deployment name. If a
selected provider key is already configured, Studio first shows only its
environment-variable name and asks whether to reuse it; accepting avoids
another secret-entry prompt while retaining a separate deployment copy. The
\texttt{--no-start} flag prepares the bundle and deployment environment without
installing or starting a user service. After a successful preparation, Studio
remembers that name. Inspect it first:

\begin{lstlisting}[style=shell]
doctor
status
\end{lstlisting}

Warnings about a missing first-run store or log and an uninstalled service are
expected at this point; a doctor failure is not. Only when service startup is
intentional and authorized, enter:

\begin{lstlisting}[style=shell]
start
status
logs
restart
status
stop
\end{lstlisting}

\texttt{start} is the first command in this sequence that changes the user
service manager. \texttt{logs} returns after showing recent output; it does not
stop the service. \texttt{stop} leaves the deployment artifacts and durable
store in place.

You may also provide a deployment name explicitly:

\begin{lstlisting}[style=shell]
status tutorial-review-studio
\end{lstlisting}

The scriptable equivalents outside Studio remain
\texttt{zippergen status tutorial-review-studio} and so on.

\begin{warningbox}{Deployment is a real operational action}
Do not enter \texttt{deploy}, \texttt{start}, or \texttt{restart} merely to
explore the menu. Read the deployment chapters first, remain on mock/fake
services, and understand the declared setup and effects. The \texttt{show full}
view exposes prompts and deployment requirements before this transition.
\end{warningbox}

\subsection{Studio and detailed CLI equivalents}

The rest of this manual intentionally exposes the lower-level operations. Use
this table to understand why those sections contain more commands.

\begin{tabularx}{\linewidth}{p{0.19\linewidth}X X}
\toprule
\textbf{Need} & \textbf{Recommended Studio path} &
\textbf{Detailed/scriptable path} \\
\midrule
Select workflow & \texttt{use} selector; remembered per project & Supply
\path{path.py:workflow} to each command. \\
Inspect & \texttt{show} selector & \texttt{zippergen show} with
\texttt{--detail}, \texttt{--communications}, or participant flags. \\
Develop & \texttt{run}; guided inputs and unique managed store &
\texttt{zippergen run} with explicit inputs, timeout, and optionally store. \\
Human decision & Inline in the same terminal & \texttt{tasks} and
\texttt{approve} from another terminal or notification adapter. \\
Continue & \texttt{resume} & Re-run against the same explicit SQLite store. \\
Specification change & \texttt{create} or
\texttt{spec refine} & Manually request a snapshot, assistant edit,
validation, views, projections, and semantic diff. \\
Deploy/operate & \texttt{deploy}, then remembered \texttt{status},
\texttt{logs}, \texttt{restart}, \texttt{stop} & Named deployment commands,
explicit options, profiles, and service-manager operations. \\
\bottomrule
\end{tabularx}

The short path should be enough to make the workflow compelling. The detailed
path that follows makes it auditable and automatable.

\section{What the reference chapters explain}
\label{sec:reference-chapters}

The opening Studio track is the recommended practical cycle. The remaining
chapters unpack its machinery so that commands can be audited, automated, and
debugged. They explain how to perform all of the following explicitly:

\begin{enumerate}
  \item Install an isolated Python environment for ZipperGen.
  \item Open the running example in Studio and discover the available
        operations without memorizing flags.
  \item Run a three-participant workflow with the built-in mock LLM.
  \item Read the global protocol, communication-only view, selected-agent view,
        and exact single-agent projections in the terminal.
  \item Run the workflow durably through SQLite and approve a human task from a
        second terminal.
  \item Ask a coding assistant to refine a copy of the workflow.
  \item Compare the old and new meanings with a semantic snapshot and diff.
  \item Create a named deployment with a source bundle and managed Python
        environment.
  \item Run that deployment in the foreground, then optionally as a supervised
        macOS or Linux user service.
  \item Diagnose, inspect, restart, reconfigure, redeploy, and stop it.
\end{enumerate}

The first pass uses the mock LLM. Studio writes managed development state below
\path{$HOME/.zippergen/workspaces} by default. The later transparent/reference
exercise chooses a separate \path{tutorial-runtime} directory so its profiles,
stores, logs, and bundles are especially easy to inspect. No API key and no
external service account are required for either mock path.

\section{Conventions and safety}

\subsection{How to read commands}

All shell commands appear in boxes such as this:

\begin{lstlisting}[style=shell]
zippergen --help
\end{lstlisting}

Do not type a leading dollar sign if another document uses one to represent the
shell prompt. Run commands one at a time and read their output before moving on.
Unless a section says otherwise, run commands from the tutorial project root---
the directory containing \path{zippergen.toml}, \path{specification.md},
\path{workflows}, and the nested \path{zippergen} framework checkout.

Text such as \placeholder{TASK-ID} is a placeholder. Replace the complete
placeholder, including angle brackets, with the value printed on your machine.

There are three places where text is entered:

\begin{tabularx}{\linewidth}{p{0.18\linewidth}p{0.27\linewidth}X}
\toprule
\textbf{Place} & \textbf{Prompt usually visible} & \textbf{Examples} \\
\midrule
Operating-system shell & \texttt{\%}, \texttt{\$}, or a directory name &
\texttt{cd}, \texttt{git}, \texttt{zippergen}, \texttt{codex} \\
ZipperGen Studio & \texttt{zippergen [workflow]>} & \texttt{show},
\texttt{models}, \texttt{run}, \texttt{deploy}, \texttt{status} \\
Coding assistant & Codex conversation prompt & The Studio-generated task;
\texttt{assistant} supplies it automatically \\
\bottomrule
\end{tabularx}

Do not enter a Studio command directly into zsh, and do not prefix a Studio
command with \texttt{zippergen} while Studio is already open.

Long shell commands use a backslash to continue on the next physical line. The
backslash must be the final character on its line---no spaces or comments may
follow it. Paste the complete listing, or run it as one line after removing the
backslashes and line breaks. Errors such as
\texttt{zsh: command not found: --input} mean the shell ended the command too
early; re-enter the complete command with intact continuations. A typographic
dash such as ``--'' copied from prose is not two ASCII hyphens; command options
must begin with two ordinary \texttt{-} characters.

\subsection{When two terminal windows are used}

Studio presents development approvals in one terminal. The later low-level and
external-adapter examples deliberately use two terminals:

\begin{itemize}
  \item \textbf{Terminal A} runs a workflow and waits.
  \item \textbf{Terminal B} checks status and approves or rejects the task.
\end{itemize}

For those later examples, both terminals must be in the project root and
must set the same \path{ZIPPERGEN_HOME} value when the guide asks for it. This
does not apply to the opening Studio path.

\subsection{Optional transparent CLI workspace}

The recommended Studio path needs no runtime export. The later low-level
exercise deliberately makes profiles, stores, logs, and bundles visible. For
that exercise, use one directory as both tutorial root and Git repository, with
runtime data in an ignored subdirectory:

\begin{lstlisting}[style=shell]
$HOME/Projects/zippergen-tutorial/  # project Git root and Codex workspace
|-- .git/
|-- zippergen.toml
|-- specification.md
|-- workflows/
|-- tests/
|-- zippergen/          # separately cloned framework, ignored by project Git
`-- tutorial-runtime/  # ignored profiles, stores, logs, and bundles
\end{lstlisting}

The \path{Projects} directory is a convenient convention, not a ZipperGen
requirement. This layout avoids putting a Git checkout directly in
\path{$HOME} and makes the entire tutorial easy to find. The lowercase,
hyphenated directory name also avoids the extra quoting and escaping that
spaces require in shell commands. Project initialization makes
\path{.gitignore} exclude both the nested framework checkout and
\path{tutorial-runtime}, so neither can be committed accidentally.
Normal Studio use and deployments use \path{$HOME/.zippergen}. The detailed
reference exercise instead sets an explicit, visible tutorial location:

\begin{lstlisting}[style=shell]
export ZIPPERGEN_HOME="$HOME/Projects/zippergen-tutorial/tutorial-runtime"
\end{lstlisting}

Run that export again in every new terminal used for the later reference
exercise. Studio does not require it. Do not change it to \path{$HOME},
\path{/}, or another broad directory.

\begin{warningbox}{Start with the mock backend}
Do not configure a paid model or live external service until the complete mock
cycle works. The mock backend returns recognizable placeholders such as
\texttt{[draft\_reply:draft]}; this is expected and proves the coordination path
without network calls or charges.
\end{warningbox}

\section{The mental model}

\subsection{A workflow is a global protocol}

Most agent frameworks distribute control logic among agents. ZipperGen instead
starts with one global protocol: who owns a value, who sends it, who receives
it, who performs an action, and who owns each decision. From that global code,
ZipperGen projects an exact local program for every participant.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[node distance=6.5mm]
  \node[zgstage, text width=22mm] (intent) {Prompt or\\human intent};
  \node[zgstage, text width=22mm, right=of intent] (source) {Python global\\workflow};
  \node[zgstage, text width=22mm, right=of source] (verify) {Validate, show,\\and diff};
  \node[zgstage, text width=22mm, right=of verify] (execute) {Durable\\execution};
  \node[zgstage, text width=22mm, right=of execute] (deploy) {Named, supervised\\deployment};
  \draw[zgflow] (intent) -- (source);
  \draw[zgflow] (source) -- (verify);
  \draw[zgflow] (verify) -- (execute);
  \draw[zgflow] (execute) -- (deploy);
\end{tikzpicture}
\caption{The prompt-to-production path. The Python workflow remains the
reviewable source of truth in the middle of the cycle.}
\label{fig:lifecycle}
\end{figure}

For a well-formed protocol, every projected send has a corresponding receive.
The code therefore serves simultaneously as executable behavior, design
documentation, and an audit surface.

A global protocol does not have to be one monolithic function. Reusable or
conceptually coherent subsequences can be given meaningful names so that the
top-level protocol stays a reasonable size. Decorate such a helper with
\texttt{@fragment}, then call it inside a \texttt{@workflow} body:

\begin{lstlisting}[style=python]
from zippergen import fragment, workflow

@fragment
def request_review(draft):
    Writer(draft) >> Reviewer(draft)
    Reviewer: approved = review(draft)
    Reviewer(approved) >> Writer(approved)

@workflow
def reviewed_answer(request: str @ Requester):
    Requester(request) >> Writer(request)
    Writer: draft = write_draft(request)
    request_review(draft)
\end{lstlisting}

A fragment is an authoring-time, reusable piece of the same global protocol.
Its statements are inlined into the surrounding workflow before validation and
projection; it is not a separately deployed workflow, participant, process, or
durable run. It can still be useful when called only once if it names a coherent
protocol stage. Use another top-level \texttt{@workflow} only when the unit
should be loaded, run, and deployed independently.

\subsection{Important words}

\begin{longtable}{p{0.20\linewidth}p{0.73\linewidth}}
\toprule
\textbf{Term} & \textbf{Meaning} \\
\midrule
\endhead
Lifeline & One sequential participant or responsibility boundary. It may
represent an agent, a person, a mailbox, or an external-system role. \\
Message & An explicit transfer of one or more values from one lifeline to
another, written with \texttt{Sender(value) >> Receiver(value)}. \\
Action & Work performed locally at one lifeline: deterministic computation,
LLM work, an external effect, human input, or runtime planning. \\
Owned decision & An \texttt{if} or \texttt{while} guard followed by
\texttt{@ Lifeline}. That lifeline must know the guard value and owns the
choice. \\
Projection & The local program generated for one lifeline from the global
protocol. \\
Workflow spec & A CLI reference such as
\path{zippergen/examples/tutorial_review.py:tutorial_review}: file path, colon, workflow
object name. It selects an existing workflow object; it does not name a
deployment. \\
Store & A SQLite file holding messages, actions, traces, human tasks, and
workflow results. It enables durable resume and replay. \\
Human task & A durable pending request created by a \texttt{@human} action. It
contains the question or rendered content, its pending/completed state, and the
eventual human decision. \\
Task ID & The stable internal identifier of a human task. A trusted local
operator may use it with \texttt{approve --task}. It identifies the task but is
not intended to be an approval credential in an untrusted external link. \\
Task token & A durable random bearer value that maps back to one human task.
External adapters can use it with \texttt{approve --token}; treat it like a
secret approval link. A token does not create another task. \\
Token channel & An arbitrary label such as \texttt{email}, \texttt{telegram},
or \texttt{cli}. It allows the same task to have a separate token for each
adapter; it is not a workflow message channel. \\
Deployment profile & Persistent non-secret settings for a named deployment. \\
Bundle & A timestamped copy of the workflow and its declared support files.
Deployments run from this snapshot, not from a file that may change halfway
through execution. \\
Managed environment & A deployment-specific Python virtual environment with
ZipperGen and declared packages installed. \\
Service manager & macOS \texttt{launchd} or Linux user \texttt{systemd}, used
to start, restart, and supervise a deployment. \\
\bottomrule
\end{longtable}

\subsection{Action kinds}

\begin{tabularx}{\linewidth}{p{0.14\linewidth}X X}
\toprule
\textbf{Decorator} & \textbf{Use it for} & \textbf{Important property} \\
\midrule
\texttt{@pure} & Deterministic local computation & No external I/O or durable
side effect. \\
\texttt{@effect} & Email, files, APIs, databases, or other external mutation &
Make it retry-safe or idempotent. \\
\texttt{@assistant} & Repository-aware work performed by Codex, Claude Code,
or a custom assistant backend & The action, Markdown instruction fingerprint,
typed inputs, owner, and result remain inspectable and durable. \\
\texttt{@llm} & Model generation or judgment & Declare prompts, parse format,
and typed outputs. \\
\texttt{@human} & Confirmation, edit, selection, acknowledgement, or input &
Creates a durable task in SQLite execution. \\
\texttt{@planner} & A model-generated allowed sub-workflow at runtime & Not the
normal mechanism for coding-assistant source generation. \\
\bottomrule
\end{tabularx}

\subsection{Coding assistants inside an executing workflow}

The Studio command \texttt{assistant codex} is a development command: it asks
an assistant to edit the current project. The \texttt{@assistant} decorator is
different. It declares repository-aware assistant work as an explicit action
inside the protocol. The owning lifeline, typed inputs, static instruction,
result, trace event, and semantic change therefore remain visible.

\begin{lstlisting}[language=Python]
from zippergen import Lifeline, assistant, workflow

Maintainer = Lifeline("Maintainer")

@assistant(
    instructions_file="prompts/update_release_notes.md",
    workspace=".",
)
def update_release_notes(change: str) -> str: ...

@workflow
def maintain(change: str @ Maintainer) -> str:
    Maintainer: report = update_release_notes(change)
    return report @ Maintainer
\end{lstlisting}

Use exactly one of \texttt{instructions=} and
\texttt{instructions\_file=}. A Markdown path is relative to the project,
fingerprinted by semantic inspection, checked by \texttt{validate}, and copied
automatically into a guided deployment bundle. Function parameters are
dynamic workflow data and are passed separately from that static instruction.

\begin{lstlisting}[language=bash]
zippergen run workflows/maintain.py:maintain \
  --assistant codex \
  --input 'change=Document the new retry policy.'
\end{lstlisting}

\texttt{--assistant claude} selects Claude Code instead. Python callers may
select a CLI or inject a custom backend:

\begin{lstlisting}[language=Python]
workflow.configure(assistant="codex")
workflow.configure(assistant_backend=my_backend)
\end{lstlisting}

A Studio run uses \texttt{run --assistant codex}; that selection is recorded so
\texttt{resume} retains it. A recorded SQLite result is replayed without
launching the assistant again.
Nevertheless, make the requested file operation safe to resume: a crash can
occur after files change but before the action result is recorded.

\section{Reference: prerequisites and installation details}
\label{sec:install}

This guide targets macOS and Linux. ZipperGen requires Python 3.11 or later.
The commands use \texttt{uv} to manage Python and the local environment. If
the opening setup already ended with a working
\texttt{zippergen --help}, do not clone or install again; use this
section only to understand or repair the setup.

\subsection{Check what is already installed}

\begin{lstlisting}[style=shell]
git --version
curl --version
python3 --version
uv --version
\end{lstlisting}

If \texttt{git} is missing, install the standard Git package for your operating
system. If \texttt{uv} is missing, use the official macOS/Linux installer:

\begin{lstlisting}[style=shell]
curl -LsSf https://astral.sh/uv/install.sh | sh
\end{lstlisting}

The installer may ask you to restart the terminal so its binary is on
\path{PATH}. The current official alternatives, including Homebrew and
\texttt{pipx}, are documented at
\url{https://docs.astral.sh/uv/getting-started/installation/}.

\subsection{Clone and install ZipperGen}

For a new checkout:

\begin{lstlisting}[style=shell]
mkdir -p "$HOME/Projects"
cd "$HOME/Projects"
mkdir zippergen-tutorial
cd zippergen-tutorial
git init
git clone https://github.com/zippergen-io/zippergen.git zippergen
uv python install 3.12
uv sync --project zippergen --python 3.12
uv tool install --force --editable ./zippergen
zippergen --help
\end{lstlisting}

\texttt{uv sync} creates or updates the nested checkout's local
\path{zippergen/.venv}. The editable tool installation exposes the short CLI
while continuing to use that source. The parent project has its own Git history
for the specification, workflows, and tests. All files remain on disk after closing a
terminal or rebooting. Normally, returning to Studio requires only:

\begin{lstlisting}[style=shell]
cd "$HOME/Projects/zippergen-tutorial"
zippergen
\end{lstlisting}

The current directory belongs to a shell session, so repeat the \texttt{cd} in
each new terminal. You do \emph{not} normally repeat \texttt{git clone},
\texttt{uv python install}, \texttt{uv sync}, \texttt{uv tool install}, or
\texttt{codex login} after a
crash or reboot. The optional low-level reference track later introduces
\texttt{ZIPPERGEN\_HOME} only to keep all runtime artifacts visible in one
tutorial directory; Studio does not need that export.

Updating the checkout is a separate, occasional operation:

\begin{lstlisting}[style=shell]
git -C zippergen status --short
git -C zippergen pull --ff-only
uv sync --project zippergen
\end{lstlisting}

If you deliberately use an existing checkout elsewhere, replace both paths
with suitable locations in that checkout, or choose another isolated runtime
directory and ensure Git ignores it.

Do not run \texttt{git pull} while you have changes you do not understand.
First inspect \texttt{git -C zippergen status --short}; commit, stash, or ask
for help when needed. Repeat \texttt{uv sync --project zippergen} after pulling
dependency changes or if the local framework environment was removed.

\subsection{Install Codex or Claude Code for Studio}
\label{sec:codex}

The nested framework checkout ships the workflow skill below, while
\path{zippergen/AGENTS.md} supplies its repository guidance:

\begin{lstlisting}[numbers=none]
zippergen/.agents/skills/zippergen-workflows/SKILL.md
\end{lstlisting}

Studio records the nested checkout in \path{zippergen.toml} and places these
exact paths in every assistant task. See the official
\href{https://learn.chatgpt.com/docs/build-skills}{Codex skill documentation}.

If the Codex CLI is not installed and Node.js/npm is available:

\begin{lstlisting}[style=shell]
npm install --global @openai/codex
codex login
\end{lstlisting}

Complete the browser sign-in once. In ordinary manual use, Codex can be started
in the parent tutorial project with:

\begin{lstlisting}[style=shell]
cd "$HOME/Projects/zippergen-tutorial"
codex
\end{lstlisting}

Codex now sees the complete application project: its manifest, canonical specification,
workflow code, tests, and nested framework checkout. Because the skill is in a
nested repository, it need not appear in the parent's \texttt{/skills} list;
the generated brief explicitly instructs Codex to read and follow it. The CLI
and IDE extension are recommended local repository surfaces; see
\url{https://learn.chatgpt.com/docs/codex/cli}.

Claude Code is an equal alternative. If it is not installed and Node.js/npm is
available, install it and start it once to complete Anthropic's authentication:

\begin{lstlisting}[style=shell]
npm install --global @anthropic-ai/claude-code
cd "$HOME/Projects/zippergen-tutorial"
claude
\end{lstlisting}

See Anthropic's
\href{https://docs.anthropic.com/en/docs/claude-code/getting-started}{Claude Code setup guide}
and
\href{https://docs.anthropic.com/en/docs/claude-code/cli-usage}{CLI reference}.

Normally, stay in Studio and enter \texttt{assistant codex} or
\texttt{assistant claude}. Studio supplies the project directory and
\path{.zippergen/current-task.md} instruction. Both tools use one-shot agent
execution, perform the task, print their report, and return to Studio
automatically instead of leaving an open assistant prompt. Codex retains the
explicit \texttt{assistant codex --interactive} mode. Commands remain subject
to each tool's configured permission rules. Neither path requires MCP. Any
MCP server already configured in the chosen tool remains optional and is not
configured or required by ZipperGen.

Coding-assistant authentication and a workflow's LLM API key are separate
matters. The mock tutorial needs no workflow API key, and ZipperGen's provider
configuration configures neither Codex nor Claude Code.

\section{The running example}

The example is \path{zippergen/examples/tutorial_review.py}. It has three lifelines:

\begin{itemize}
  \item \textbf{Requester} initially owns a request and the maximum number of
        revision attempts, and receives the answer.
  \item \textbf{Writer} drafts a reply and revises it after rejection.
  \item \textbf{Reviewer} uses a separate LLM action to assess every draft,
        presents the draft and notes to a human authority, and owns the bounded
        retry loop.
\end{itemize}

The protocol is:

\begin{enumerate}
  \item Requester validates \texttt{max\_retries}, sends the request to Writer,
        and sends the retry budget to Reviewer.
  \item Writer uses an LLM action to create a draft.
  \item Writer sends the draft to Reviewer.
  \item Reviewer uses an LLM action to identify concerns, then presents both
        the draft and concerns in a human confirmation action.
  \item If Reviewer approves, the draft goes to Requester.
  \item If Reviewer rejects while revision attempts remain, Reviewer returns
        the draft to Writer; Writer revises it and presents the new draft to
        Reviewer again.
  \item If the retry budget is exhausted without approval, Requester receives
        an explicit failure notice instead of an unapproved draft.
\end{enumerate}

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[node distance=17mm and 12mm]
  \node[zgstage, text width=23mm] (draft) {Writer creates\\initial draft};
  \node[zgstage, text width=24mm, right=of draft] (assess)
    {Reviewer LLM\\assesses draft};
  \node[zgplain, diamond, aspect=2.0, text width=20mm, right=of assess]
    (review) {Reviewer\\approves?};
  \node[zgstage, text width=23mm, right=of review] (deliver) {Deliver approved\\draft};
  \node[zgplain, diamond, aspect=2.0, text width=20mm, below=of review]
    (budget) {Retries\\remain?};
  \node[zgstage, text width=24mm, left=of budget] (revise) {Writer revises\\draft};
  \node[zgplain, text width=27mm, draw=ZipperRed, fill=ZipperRed!8,
    right=of budget] (failure) {Return explicit\\failure notice};
  \draw[zgflow] (draft) -- (assess);
  \draw[zgflow] (assess) -- (review);
  \draw[zgflow] (review) -- node[zglabel, above]{yes} (deliver);
  \draw[zgflow] (review) -- node[zglabel, right]{no} (budget);
  \draw[zgflow] (budget) -- node[zglabel, above]{yes} (revise);
  \draw[zgflow] (revise) to[bend left=28]
    node[zglabel, left]{assess again} (assess);
  \draw[zgdashed] (budget) -- node[zglabel, above]{no} (failure);
\end{tikzpicture}
\caption{A rejection produces another reviewed draft only while the bounded
retry budget remains. No unapproved draft is silently released.}
\label{fig:review-loop}
\end{figure}

\begin{notebox}{What \texttt{max\_retries} counts}
The initial draft is not a retry. With \texttt{max\_retries=2}, Writer may make
at most two revisions after rejection, so Reviewer sees at most three human
tasks: the initial draft and two revised drafts.
\end{notebox}

The complete, executable source is included in Appendix~\ref{app:source}.

\subsection{Why the deployment declaration matters}

The module also defines \texttt{zippergen\_deployment}. It tells the guided
deployer to collect:

\begin{itemize}
  \item the LLM setting, defaulting to \texttt{mock};
  \item the request text as a workflow input;
  \item the maximum number of revisions as a second workflow input, defaulting
        to \texttt{2};
  \item an OpenAI API key only if the default or any per-lifeline model starts
        with \texttt{openai}; and
  \item an Anthropic API key only if the default or any per-lifeline model
        starts with \texttt{anthropic} or \texttt{claude}.
\end{itemize}

The declaration is data-only. A coding assistant can generate it, and
\texttt{show --detail full}, \texttt{validate}, \texttt{doctor}, and
\texttt{deploy} can inspect it without bespoke shell scripts.

\section{Optional reference: one-process in-memory execution}

Studio's durable \texttt{run} is the recommended development command. This
short direct run is useful only for understanding the simpler in-memory
backend or debugging the Python module itself. Run the example directly:

\begin{lstlisting}[style=shell]
uv run --project zippergen python zippergen/examples/tutorial_review.py
\end{lstlisting}

Expected interaction:

\begin{lstlisting}[style=shell]
Draft:
[draft_reply:draft]

Automated reviewer notes:
[assess_draft:concerns]

Approve this draft? [y/n]: y
Result: [draft_reply:draft]
\end{lstlisting}

Run it again and enter \texttt{n}. Writer produces
\texttt{[revise\_reply:draft]}, Reviewer produces a new
\texttt{[assess\_draft:concerns]}, and the human is asked again; enter
\texttt{y} to accept that revision. If you keep entering \texttt{n}, the program stops after
two revision attempts and returns an explicit failure notice. This direct
program uses in-memory execution and an in-terminal human backend. It is
convenient for the first smoke test, but it does not demonstrate durable
restart behavior.

Internally, this memory backend runs entirely inside the one Python process
started by \texttt{uv run --project zippergen python}. It creates one Python thread for each
lifeline---Requester, Writer, and Reviewer in this example---and uses
thread-safe FIFO queues, keyed by sender, receiver, and channel, for their
messages. It does not create a separate operating-system process for each
lifeline. When the program exits or crashes, the threads, queues, messages, and
progress disappear together. The later SQLite backend still runs local
lifelines in threads, but records communication and execution progress in the
database so it can recover them after a restart.

\section{Optional reference: inspect with scriptable CLI commands}

Set the workflow spec once in your head:
\path{zippergen/examples/tutorial_review.py:tutorial_review}. Read it as follows:

\begin{itemize}
  \item Before the colon, \path{zippergen/examples/tutorial_review.py} is the Python file
  to load.
  \item After the colon, \texttt{tutorial\_review} is the existing Python
  object to select. It comes from
  \texttt{@workflow def tutorial\_review(...)} inside that file.
\end{itemize}

The suffix does not assign a new name; it locates the workflow object. The
deployment name is separate: this example declares \texttt{tutorial-review}
in \texttt{zippergen\_deployment}, and deployment commands may supply it with
\texttt{--name tutorial-review}. The underscore in the Python object name and
the hyphen in the deployment name are intentional. The validation JSON makes
the distinction visible as \texttt{"workflow": "tutorial\_review"} versus
\texttt{"deployment": \{"name": "tutorial-review", ...\}}.

The CLI currently expects the workflow spec as a positional argument, so the
following commands repeat it for clarity.

\subsection{Validate before running or deploying}

\begin{lstlisting}[style=shell]
zippergen validate zippergen/examples/tutorial_review.py:tutorial_review
\end{lstlisting}

Successful validation reports three unique lifelines, successful projections
for Requester, Writer, and Reviewer, a valid deployment declaration, and
a successful \emph{canonical rendering} check. Here, canonical rendering means
that ZipperGen can turn the loaded workflow into its standard, normalized code
views: the global protocol and the generated local view for each participant.
It does not execute or deploy the workflow. Treat a validation failure as a
blocker.

For machine-readable output:

\begin{lstlisting}[style=shell]
zippergen validate zippergen/examples/tutorial_review.py:tutorial_review --json
\end{lstlisting}

\subsection{Global detail levels and the communication-only view}

The plain \texttt{show} command uses \texttt{--detail protocol} by default, so
the first two protocol commands below are equivalent:

\begin{lstlisting}[style=shell]
zippergen show zippergen/examples/tutorial_review.py:tutorial_review
zippergen show zippergen/examples/tutorial_review.py:tutorial_review --detail protocol
zippergen show zippergen/examples/tutorial_review.py:tutorial_review --detail overview
zippergen show zippergen/examples/tutorial_review.py:tutorial_review \
  --detail protocol --communications
zippergen show zippergen/examples/tutorial_review.py:tutorial_review --detail actions
zippergen show zippergen/examples/tutorial_review.py:tutorial_review --detail full
\end{lstlisting}

The four detail levels are:

\begin{itemize}
  \item \texttt{overview}: names the workflow, lifelines, inputs, outputs,
  actions, and deployment counts, but replaces the workflow body with an
  ellipsis.
  \item \texttt{protocol}: shows the complete global workflow body, including
  communications, decisions, loops, parallel regions, and action calls. This
  is the default.
  \item \texttt{actions}: adds the action declarations and signatures; detailed
  LLM and planner prompt bodies remain hidden.
  \item \texttt{full}: adds prompts, available implementations, and the
  normalized deployment declaration.
\end{itemize}

The separate \texttt{--communications} filter keeps messages, decisions, and
the return value but omits local action calls. It can be combined with a detail
level when useful. Because \texttt{protocol} is the default,
\texttt{--communications} alone is shorthand for
\texttt{--detail protocol --communications}; this guide spells out the full
form where the view is introduced.

\subsection{One participant or a selected group}

Exact single-agent projections:

\begin{lstlisting}[style=shell]
zippergen show zippergen/examples/tutorial_review.py:tutorial_review --agent Requester
zippergen show zippergen/examples/tutorial_review.py:tutorial_review --agent Writer
zippergen show zippergen/examples/tutorial_review.py:tutorial_review --agent Reviewer
\end{lstlisting}

A focused multi-agent view keeps hidden peers as explicit external boundaries:

\begin{lstlisting}[style=shell]
zippergen show zippergen/examples/tutorial_review.py:tutorial_review \
  --agents Writer,Reviewer
\end{lstlisting}

Do not confuse these modes: \texttt{--agent Reviewer} is the exact formal local
projection; \texttt{--agents Writer,Reviewer} is a global focus view.

Structured output for tools and coding assistants is also available:

\begin{lstlisting}[style=shell]
zippergen show zippergen/examples/tutorial_review.py:tutorial_review --format json
\end{lstlisting}

\section{Optional reference: explicit durable store and two terminals}

Studio already provides this behavior with a managed store and inline human
decision. This lower-level pass is useful for automation, notification
adapters, or understanding the exact SQLite/task commands. The process in
Terminal A waits without losing its state; Terminal B finds and completes the
human task.

\subsection{Task, task ID, and token}

The \emph{task} is the pending work itself: in this example, the rendered draft
and the request for a human decision. The \emph{task ID} is its stable internal
name. A trusted operator working locally can use that ID to select the task. A
\emph{token} is a separate, random bearer credential that an external adapter
can use instead of trusting a raw task ID. Protect a token like an approval
link. Both point to the same task; neither is the approval or rejection
decision.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[node distance=18mm and 20mm]
  \node[zgplain, text width=23mm] (cli) {Trusted local\\CLI operator};
  \node[zgstore, text width=31mm, right=of cli] (task) {One human task\\rendered draft\\state: pending};
  \node[zgplain, text width=25mm, right=of task] (decision) {Decision\\\texttt{--yes} or\\\texttt{--no}};
  \node[zgplain, text width=23mm, below=of cli] (adapter) {External adapter\\email, Telegram,\\or Slack};
  \node[zgplain, text width=31mm, draw=ZipperGold,
    fill=ZipperGold!10, below=of task] (token) {Token record\\task + channel\\+ random token};
  \draw[zgflow] (cli) -- node[zglabel, above]{\texttt{--task TASK-ID}} (task);
  \draw[zgflow] (adapter) -- node[zglabel, above]{presents \texttt{TOKEN}} (token);
  \draw[zgflow] (token) -- node[zglabel, right]{maps to} (task);
  \draw[zgflow] (task) -- node[zglabel, above]{resolves} (decision);
\end{tikzpicture}
\caption{The human task is the pending work. A task ID is a local locator; a
token is an adapter-facing credential that resolves to that same task. The
separate \texttt{--yes}/\texttt{--no} flag supplies the decision.}
\label{fig:task-token}
\end{figure}

\subsection{Terminal A: start the workflow}

From the project root:

\begin{lstlisting}[style=shell]
export ZIPPERGEN_HOME="$HOME/Projects/zippergen-tutorial/tutorial-runtime"
STORE="$ZIPPERGEN_HOME/runs/manual-review.sqlite"

zippergen run zippergen/examples/tutorial_review.py:tutorial_review \
  --llm mock \
  --input 'request=Explain why the sky is blue in two sentences.' \
  --input 'max_retries=2' \
  --store "$STORE" \
  --timeout 0
\end{lstlisting}

Each \texttt{--input} occurrence supplies one \texttt{name=value} pair. The
names match the parameters in
\texttt{tutorial\_review(request, max\_retries)}. ZipperGen parses
\texttt{max\_retries=2} as an integer. For many inputs, the two flags may be
replaced by one JSON object:

\begin{lstlisting}[style=shell]
--input-json '{"request":"Explain why the sky is blue.","max_retries":2}'
\end{lstlisting}

\texttt{STORE} is merely a short shell variable chosen by this tutorial, not a
special ZipperGen setting. The shell replaces \texttt{\$STORE} with the full
SQLite path before running each command. Terminal A prints that path and then
waits. Leave it running.
\texttt{--timeout 0} means no execution deadline; use this only when waiting is
intentional.

\subsection{Terminal B: inspect status and tasks}

Open a second terminal, enter the repository, and repeat both variable
assignments because a new terminal does not inherit them:

\begin{lstlisting}[style=shell]
cd "$HOME/Projects/zippergen-tutorial"
export ZIPPERGEN_HOME="$HOME/Projects/zippergen-tutorial/tutorial-runtime"
STORE="$ZIPPERGEN_HOME/runs/manual-review.sqlite"

zippergen status \
  --store "$STORE"

zippergen tasks \
  --store "$STORE"
\end{lstlisting}

Status should say \texttt{waiting}. The tasks command prints the rendered draft
and a task identifier. Copy the task identifier. Choose exactly one of the
following decisions; once the task is decided, do not run the other command for
the same task.

Approve it:

\begin{lstlisting}[style=shell]
zippergen approve \
  --store "$STORE" \
  --task <TASK-ID> --yes
\end{lstlisting}

Or, to exercise the rejection branch instead, reject it explicitly:

\begin{lstlisting}[style=shell]
zippergen approve \
  --store "$STORE" \
  --task <TASK-ID> --no
\end{lstlisting}

After rejection, Terminal A asks Writer for a revision and then waits on a
\emph{new} human task for the revised draft. List the pending tasks again, copy
the new task ID, and make the next decision:

\begin{lstlisting}[style=shell]
zippergen tasks --store "$STORE"

zippergen approve \
  --store "$STORE" \
  --task <NEW-TASK-ID> --yes
\end{lstlisting}

Every presented draft is a distinct task with its own task ID. Approving any
draft finishes successfully. Rejecting the initial draft and both permitted
revisions exhausts \texttt{max\_retries=2}; Terminal A then returns the explicit
failure notice and exits. Do not reuse an already-completed task ID.

The task ID is sufficient for this trusted local CLI exercise. External
adapters such as email, Telegram, or Slack can instead use durable tokens. To
display one, rerun the task listing as follows:

\begin{lstlisting}[style=shell]
zippergen tasks \
  --store "$STORE" \
  --tokens --channel cli
\end{lstlisting}

Here \texttt{cli} is only a descriptive token-namespace label; it is not a
workflow communication channel. Different adapter labels may have different
tokens for the same task. A token replaces the task ID, while \texttt{--yes} or
\texttt{--no} still determines the decision. For example, token-based rejection
is:

\begin{lstlisting}[style=shell]
zippergen approve \
  --store "$STORE" \
  --token <TOKEN> --no
\end{lstlisting}

After a draft is approved or the retry budget is exhausted, Terminal A prints
a JSON result and exits. Confirm completion and view the trace:

\begin{lstlisting}[style=shell]
zippergen status \
  --store "$STORE"

zippergen trace \
  --store "$STORE" --tail 30

zippergen tasks \
  --store "$STORE" --all
\end{lstlisting}

\begin{notebox}{Why rerunning may finish immediately}
The SQLite store records a completed result for this workflow. Running the same
workflow against the same store replays committed work instead of repeating it.
Point \texttt{STORE} at a new, explicitly named file for a genuinely new
tutorial run, for example
\texttt{STORE="\$ZIPPERGEN\_HOME/runs/manual-review-2.sqlite"}. Never delete a
production store merely to force a rerun.
\end{notebox}

\subsection{Recover after a terminal or computer crash}

The foreground process stops, but the SQLite file remains. Open a new terminal,
restore the working directory and tutorial variables, inspect the store, and
rerun the same command with the same \texttt{STORE} value:

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[node distance=19mm]
  \node[zgstage, text width=27mm] (first) {First workflow\\process};
  \node[zgstore, text width=31mm, right=of first] (store) {SQLite store\\committed events\\and pending task};
  \node[zgstage, text width=29mm, right=of store] (second) {New workflow\\process};
  \node[zgplain, text width=27mm, draw=ZipperRed,
    fill=ZipperRed!8, below=13mm of first] (crash) {Terminal closes\\or computer crashes};
  \draw[zgflow] (first) -- node[zglabel, above]{writes} (store);
  \draw[zgdashed] (first) -- node[zglabel, right]{process stops} (crash);
  \draw[zgflow] (store) -- node[zglabel, above]{same \texttt{--store}} (second);
  \node[font=\scriptsize, text=ZipperGreen, align=center, below=7mm of second]
    {replay committed work\\then resume};
\end{tikzpicture}
\caption{A crash removes the process, not the durable history. Starting a new
process with the same SQLite store replays committed work and resumes.}
\label{fig:durable-resume}
\end{figure}

\begin{lstlisting}[style=shell]
cd "$HOME/Projects/zippergen-tutorial"
export ZIPPERGEN_HOME="$HOME/Projects/zippergen-tutorial/tutorial-runtime"
STORE="$ZIPPERGEN_HOME/runs/manual-review.sqlite"

zippergen status \
  --store "$STORE"

zippergen run zippergen/examples/tutorial_review.py:tutorial_review \
  --llm mock \
  --input 'request=Explain why the sky is blue in two sentences.' \
  --input 'max_retries=2' \
  --store "$STORE" \
  --timeout 0
\end{lstlisting}

The runner replays committed events and continues from the durable state. You
do not recreate the checkout or virtual environment, reinstall packages, or
activate the virtual environment: \texttt{uv run} selects it automatically. An
in-memory run cannot be recovered this way because it has no SQLite record.

\section{Develop and refine with a coding assistant}

This section works on a copy so the canonical example remains available for
comparison and deployment.

\subsection{Create a branch and working copy}

\begin{lstlisting}[style=shell]
git switch -c tutorial/workflow-cycle
mkdir -p workbench
cp zippergen/examples/tutorial_review.py workbench/tutorial_review.py
git status --short
\end{lstlisting}

The \texttt{-c} form creates a new branch and therefore runs only once. If a
branch of that name already exists from your own earlier attempt, use
\texttt{git switch tutorial/workflow-cycle} instead. If
\path{workbench/tutorial_review.py} already exists, inspect it before deciding
whether to reuse it; do not overwrite work you do not understand.

Save a semantic baseline before any edit:

\begin{lstlisting}[style=shell]
export ZIPPERGEN_HOME="$HOME/Projects/zippergen-tutorial/tutorial-runtime"
mkdir -p "$ZIPPERGEN_HOME/snapshots"
SNAPSHOT="$ZIPPERGEN_HOME/snapshots/tutorial-before-$(date +%Y%m%d-%H%M%S).json"

zippergen snapshot \
  workbench/tutorial_review.py:tutorial_review \
  -o "$SNAPSHOT"
echo "$SNAPSHOT"
\end{lstlisting}

The timestamp makes the baseline unique, so rerunning the exercise cannot
silently replace the meaning you intended to preserve. Copy the absolute path
printed by \texttt{echo}; it is the value used below.

\subsection{Use a precise refinement prompt}

In Studio, select the workbench copy and open the one automatically named
pending refinement:

\begin{lstlisting}[style=shell]
use workbench/tutorial_review.py:tutorial_review
spec refine
\end{lstlisting}

Paste the following text into the editor, save it, and exit:

\begin{lstlisting}[numbers=none]
# Add compliance review after exhaustion

If Reviewer exhausts the max_retries budget without approving a draft, a
separate Compliance human participant reviews the last draft. If Compliance
approves, release that draft to Requester; otherwise send the existing explicit
failure notice. Keep all Reviewer-approved paths and the retry bound unchanged.
\end{lstlisting}

Studio stores the text at \path{.zippergen/pending-refinement.md}, records its
own semantic baseline, and generates the complete task. No prompt filename or
baseline placeholder is required. Inspect and execute it:

\begin{lstlisting}[style=shell]
spec pending
task
assistant
\end{lstlisting}

The generated task supplies the nested skill, canonical specification, pending
change, target, semantic baseline, verification, and explicit no-deployment
boundary. Studio starts Codex in the tutorial root with that fixed task.

\subsection{Independently verify the assistant's work}

Even if the assistant reports success, run these yourself. If this is a new
terminal, first restore \texttt{SNAPSHOT} using the absolute path printed
earlier:

\begin{lstlisting}[style=shell]
cd "$HOME/Projects/zippergen-tutorial"
SNAPSHOT='<ABSOLUTE-SEMANTIC-BASELINE-PATH>'
\end{lstlisting}

Replace the placeholder and keep the single quotes. Then run:

\begin{lstlisting}[style=shell]
git diff -- workbench tests
git status --short

zippergen validate workbench/tutorial_review.py:tutorial_review
zippergen show workbench/tutorial_review.py:tutorial_review \
  --detail protocol --communications
zippergen show workbench/tutorial_review.py:tutorial_review \
  --agent Reviewer
zippergen show workbench/tutorial_review.py:tutorial_review \
  --agent Compliance

zippergen diff \
  "$SNAPSHOT" \
  workbench/tutorial_review.py:tutorial_review
\end{lstlisting}

The diff should show a Compliance lifeline, a new human action, new messages and
control structure on the retry-exhaustion path, and the deployment file-path
change. It should not show changes to Reviewer-approved paths or to the retry
bound. The semantic diff ignores comments and formatting but detects workflow
renames, action implementation changes, participant ownership,
message/control context, execution order, parallel structure, and deployment
requirements.

Run focused tests supplied by the assistant, then the full suite:

\begin{lstlisting}[style=shell]
uv run --project zippergen python -m pytest -q
uvx pyright
\end{lstlisting}

Commit only after reviewing the change:

\begin{lstlisting}[style=shell]
git add workbench/tutorial_review.py
git add tests/<ASSISTANT-CREATED-TEST-FILE>.py
git add specification.md
git diff --cached
git commit -m "Add compliance review tutorial variant"
\end{lstlisting}

Pushing is optional and depends on whether the branch belongs on your remote:

\begin{lstlisting}[style=shell]
git push -u origin tutorial/workflow-cycle
\end{lstlisting}

\subsection{Specification template for a workflow created from scratch}

Enter \texttt{create} and use this structure in the automatically opened
\path{specification.md}:

\begin{lstlisting}[numbers=none]
# Workflow title

- Purpose and successful outcome: ...
- Participants and responsibilities: ...
- Inputs and initial owners: ...
- Messages and data crossing boundaries: ...
- LLM, deterministic, external-effect, and human actions: ...
- Decision and loop owners: ...
- Parallel work: ...
- Final outputs and owners: ...
- External services, credentials, and deployment constraints: ...
- Safety/idempotency requirements: ...
\end{lstlisting}

Later behavioral changes go into the single pending document opened by
\texttt{spec refine}. The assistant integrates that change into this clean
canonical specification; after human review, Studio reconciles the pending
document. Git, rather than a pile of active prompt fragments, preserves
accepted history.

\section{Create a guided named deployment}

Return to the canonical example for a deterministic deployment exercise. Set
the tutorial runtime directory in the current terminal:

\begin{lstlisting}[style=shell]
export ZIPPERGEN_HOME="$HOME/Projects/zippergen-tutorial/tutorial-runtime"
\end{lstlisting}

Create the bundle, managed environment, profile, and service definitions, but
do not start a background service yet:

\begin{lstlisting}[style=shell]
zippergen deploy zippergen/examples/tutorial_review.py:tutorial_review \
  --name tutorial-review \
  --llm mock \
  --input 'request=Explain deployment in two simple sentences.' \
  --input 'max_retries=2' \
  --yes \
  --no-start
\end{lstlisting}

The default path performs these steps:

\begin{enumerate}
  \item load the source workflow and deployment declaration;
  \item collect defaults and command-line overrides;
  \item copy declared source files into a timestamped bundle;
  \item create a deployment-specific Python environment;
  \item install ZipperGen and declared packages;
  \item write the profile, run script, launchd plist, and systemd unit template;
  \item run doctor checks; and
  \item stop before service startup because of \texttt{--no-start}.
\end{enumerate}

\texttt{--yes} accepts declared defaults non-interactively. Omit it to see the
guided questions. It never means ``approve arbitrary live effects''; it only
controls configuration prompting.

\subsection{Inspect the generated artifacts}

\begin{lstlisting}[style=shell]
find "$ZIPPERGEN_HOME" -maxdepth 5 -type f -print | sort
zippergen doctor tutorial-review
\end{lstlisting}

Before the first run, doctor normally warns that the SQLite store and log do not
exist and that the service is not installed. Those warnings are expected;
failures are not.

\begin{longtable}{p{0.38\linewidth}p{0.55\linewidth}}
\toprule
\textbf{Path below \texttt{ZIPPERGEN\_HOME}} & \textbf{Purpose} \\
\midrule
\endhead
\path{deployments/tutorial-review.json} & Public deployment profile. \\
\path{deployments/tutorial-review.secrets.json} & Mode-0600 environment
secrets, created when required. It is empty for the mock tutorial. \\
\path{deployments/tutorial-review.sh} & Stable launch script. \\
\path{deployments/io.zippergen.tutorial-review.plist} & Generated macOS
launchd definition. \\
\path{deployments/zippergen-tutorial-review.service} & Generated Linux systemd
user-unit template. \\
\path{environments/tutorial-review/} & Managed Python environment. \\
\path{apps/tutorial-review/<timestamp>/} & Timestamped source bundle for this
deployment revision. \\
\path{runs/tutorial-review.sqlite} & Durable execution store. \\
\path{logs/tutorial-review.log} & Supervised-service standard output/error. \\
\bottomrule
\end{longtable}

\section{Run the named deployment in the foreground}

Foreground execution is the easiest way to test the deployment profile and
bundle before involving a service manager.

\subsection{Terminal A}

\begin{lstlisting}[style=shell]
cd "$HOME/Projects/zippergen-tutorial"
export ZIPPERGEN_HOME="$HOME/Projects/zippergen-tutorial/tutorial-runtime"
zippergen run-deployment tutorial-review
\end{lstlisting}

\subsection{Terminal B}

\begin{lstlisting}[style=shell]
cd "$HOME/Projects/zippergen-tutorial"
export ZIPPERGEN_HOME="$HOME/Projects/zippergen-tutorial/tutorial-runtime"

zippergen status tutorial-review
zippergen tasks \
  --store "$ZIPPERGEN_HOME/runs/tutorial-review.sqlite"

zippergen approve \
  --store "$ZIPPERGEN_HOME/runs/tutorial-review.sqlite" \
  --task <TASK-ID> --yes
\end{lstlisting}

Terminal A prints the result and exits. Then run:

\begin{lstlisting}[style=shell]
zippergen status tutorial-review
zippergen trace tutorial-review --tail 30
zippergen doctor tutorial-review
\end{lstlisting}

This proves that the bundled code, managed Python, persisted inputs, mock LLM,
SQLite store, and approval backend work together.

\section{Optional: run as a supervised user service}

The example is finite, while supervised deployments are primarily intended for
long-running watchers and servers. To test supervision safely, use a second
deployment name and stop it immediately after completion.

\subsection{Prepare a fresh service deployment}

\begin{lstlisting}[style=shell]
export ZIPPERGEN_HOME="$HOME/Projects/zippergen-tutorial/tutorial-runtime"

zippergen deploy zippergen/examples/tutorial_review.py:tutorial_review \
  --name tutorial-review-service \
  --llm mock \
  --input 'request=Demonstrate durable restart in two sentences.' \
  --input 'max_retries=2' \
  --yes --no-start

zippergen start tutorial-review-service --dry-run
\end{lstlisting}

Read the dry-run commands. On macOS they use \texttt{launchctl}; on Linux they
use \texttt{systemctl --user}. If they look appropriate, start and enable the
service:

\begin{lstlisting}[style=shell]
zippergen start tutorial-review-service --enable
zippergen doctor tutorial-review-service
zippergen status tutorial-review-service
\end{lstlisting}

The status should become \texttt{waiting}. Test durability by restarting while
the human task is still pending:

\begin{lstlisting}[style=shell]
zippergen restart tutorial-review-service
zippergen status tutorial-review-service
\end{lstlisting}

The same task remains pending because it lives in SQLite, not only in process
memory.

Because this service was started with \texttt{--enable}, its generated launcher
records \path{ZIPPERGEN_HOME}. After a full reboot, the user service manager
normally starts it again when the user session starts: after graphical login on
macOS, or when the user systemd manager starts on Linux. No interactive shell
export is needed by the service itself. You still repeat the \texttt{cd} and
\texttt{export} commands before using ZipperGen CLI commands in a new terminal.

\subsection{Watch logs and approve}

In another terminal you may follow the log:

\begin{lstlisting}[style=shell]
export ZIPPERGEN_HOME="$HOME/Projects/zippergen-tutorial/tutorial-runtime"
zippergen logs tutorial-review-service --follow
\end{lstlisting}

Press Control-C to stop following the log; that does not stop the service.
List and approve the task:

\begin{lstlisting}[style=shell]
zippergen tasks \
  --store "$ZIPPERGEN_HOME/runs/tutorial-review-service.sqlite"

zippergen approve \
  --store "$ZIPPERGEN_HOME/runs/tutorial-review-service.sqlite" \
  --task <TASK-ID> --yes

zippergen status tutorial-review-service
zippergen stop tutorial-review-service
\end{lstlisting}

Stop promptly after the finite workflow completes. The generated supervisor is
configured to keep a deployment alive, which is correct for polling workflows
such as call intake but unnecessary for this finite tutorial.

If the machine does not support launchd or systemd user services, keep using:

\begin{lstlisting}[style=shell]
zippergen run-deployment tutorial-review-service
\end{lstlisting}

under your preferred terminal multiplexer or process supervisor.

\section{Update, reconfigure, and redeploy}

\subsection{Change persistent configuration}

For a long-running deployment, update declared fields and restart in one step:

\begin{lstlisting}[style=shell]
zippergen configure tutorial-review-service \
  --input 'request=A new configured request.' \
  --input 'max_retries=3' \
  --yes --restart
\end{lstlisting}

For this finite example, the existing store already contains a completed
result. Changing its input does not erase durable history or create a new job.
Use a new deployment name/store for a new finite execution. For a real watcher,
\texttt{configure --restart} is the normal pattern.

\subsection{Redeploy changed source}

After editing and validating the source, redeploy an existing named deployment:

\begin{lstlisting}[style=shell]
zippergen validate zippergen/examples/tutorial_review.py:tutorial_review
zippergen deploy tutorial-review-service --yes
\end{lstlisting}

This creates a new timestamped source bundle and starts the named deployment.
The stable SQLite store is retained, so already committed durable work is not
repeated.

To deploy the assistant-modified workbench version as a separate test:

\begin{lstlisting}[style=shell]
zippergen deploy workbench/tutorial_review.py:tutorial_review \
  --name tutorial-review-v2 \
  --llm mock \
  --input 'request=Test the compliance version.' \
  --input 'max_retries=2' \
  --yes --no-start
\end{lstlisting}

\subsection{Switch lifelines to real models only after mock success}

Replace \placeholder{OPENAI-MODEL} and \placeholder{ANTHROPIC-MODEL} with
identifiers available to your accounts. Keep the inherited default on mock and
override the two LLM-active lifelines. Omit \texttt{--yes} so the guided
command asks for both enabled secrets without putting either in shell history:

\begin{lstlisting}[style=shell]
zippergen configure tutorial-review-service \
  --llm mock \
  --llm-for 'Writer=openai:<OPENAI-MODEL>' \
  --llm-for 'Reviewer=claude:<ANTHROPIC-MODEL>' \
  --restart
\end{lstlisting}

The conditional OpenAI and Anthropic key fields now become active. ZipperGen
stores the values in the deployment's private mode-0600 secrets file, loads
them before importing the bundled workflow, and does not copy them into the
public profile or service definition. Use
\texttt{--llm-for Writer=inherit} to remove an existing override. If the
service is intentionally stopped, omit \texttt{--restart}; test later with
\texttt{run-deployment} or start it explicitly.

\begin{warningbox}{Never paste secrets into source or specifications}
Do not commit API keys, include them in a coding-assistant prompt, pass them via
\texttt{--set} on a shared shell history, or add them to a normal deployment
profile. Use the guided secret prompt or a securely managed environment.
\end{warningbox}

\section{Troubleshooting}

\subsection*{\texttt{uv: command not found}}

Restart the terminal after installing uv, then run \texttt{uv --version}. If it
still fails, follow the PATH instructions printed by the official installer.

\subsection*{\texttt{studio} is reported as an invalid top-level command}

The nested framework checkout or editable tool is older than this guide, or
the terminal is in another project. From the tutorial project root, inspect
before updating:

\begin{lstlisting}[style=shell]
pwd
git -C zippergen status --short
git -C zippergen branch --show-current
git -C zippergen remote -v
uv sync --project zippergen --python 3.12
uv tool install --force --editable ./zippergen
zippergen --help
\end{lstlisting}

If the worktree is clean and this is the intended remote/branch, use
\texttt{git -C zippergen pull --ff-only} and rerun the sync and editable tool
installation. ``Already up to date''
only means Git found no newer commit on the configured upstream; it does not
prove that the terminal is in the intended checkout or branch.

\subsection*{The shell reports \texttt{models} as an invalid command}

This is expected if you entered \texttt{zippergen models} in the
operating-system shell. \texttt{models} is a Studio command, not a top-level CLI
subcommand. Start Studio from the Git root, then enter it at the Studio prompt:

\begin{lstlisting}[style=shell]
zippergen
# The next line is entered only after zippergen [workflow]> appears:
models show
\end{lstlisting}

If no workflow is selected, enter \texttt{use} first.

\subsection*{Studio says the specification is empty}

Enter \texttt{spec path} to see Studio's fixed path, then \texttt{spec edit} to
open it again. Add at least one real requirement, save, and leave the editor.
The file must be valid nonempty UTF-8 text. No prompt directory, ID, or custom
filename is required. If importing an existing document with the advanced
\texttt{create --file PATH} form, Studio reports the resolved absolute path
and distinguishes a missing file, directory, empty file, and invalid UTF-8.

\subsection*{The shell says \texttt{command not found: --input} or \texttt{--store}}

The preceding multiline command ended early. Re-enter the entire command and
ensure every continuation backslash is the final character on its line. This
same mistake can cause ZipperGen to report a missing workflow input because
only the first part of the intended command actually ran.

\subsection*{\texttt{models show} lists no Writer or Reviewer}

Run \texttt{current} to confirm the selected workflow, then
\texttt{show actions}. Studio lists a lifeline in \texttt{models} only when that
lifeline contains an actual \texttt{@llm} action. A pure, human, or message-only
lifeline does not need a model. If the generated workflow was meant to contain
those actions, treat their absence as a generation/validation problem rather
than assigning a model to an unrelated participant.

\subsection*{A real model run reports an authentication or model error}

Use \texttt{models show} to inspect provider prefixes and model identifiers,
then use \texttt{models check} for a fresh, read-only availability check.
Confirm that the account key belongs to that provider and that the named model
is enabled for the account. Do not print the key while diagnosing. To continue
the tutorial without network access or charges, enter
\texttt{models reset all} and start a fresh mock \texttt{run}.

\subsection*{The workflow file cannot be found}

Check the current directory and spec:

\begin{lstlisting}[style=shell]
pwd
ls zippergen/examples/tutorial_review.py
zippergen validate zippergen/examples/tutorial_review.py:tutorial_review
\end{lstlisting}

The text after the colon is the Python workflow object name, not the deployment
name.

\subsection*{Git asks for \texttt{user.name} or \texttt{user.email}}

Git cannot create the local commit until an author identity is configured.
Follow your organization or hosting provider's identity policy. To configure
only this tutorial checkout, run from its root:

\begin{lstlisting}[style=shell]
git config user.name "Your Name"
git config user.email "your-address@example.com"
git commit -m "Add reviewed answer workflow"
\end{lstlisting}

Replace both example values. Repository-local configuration avoids changing
other checkouts on the computer.

\subsection*{Validation or projection fails}

Do not deploy. Read the first failing check, inspect the global view and the
affected local projection, then ask the coding assistant to diagnose the error.
Common causes are missing messages, a decision owner that does not possess the
guard value, an output that is not present at the declared lifeline, or a
parallel data dependency.

\subsection*{No human task appears}

Confirm that Terminal A is still running and both commands use the same store:

\begin{lstlisting}[style=shell]
zippergen status --store /exact/path/to/store.sqlite
zippergen tasks --store /exact/path/to/store.sqlite --all
\end{lstlisting}

If status is \texttt{done}, that store already contains a result. Use a new
store path for a new finite run.

\subsection*{The service will not start}

\begin{lstlisting}[style=shell]
zippergen doctor tutorial-review-service
zippergen start tutorial-review-service --dry-run
zippergen logs tutorial-review-service --tail 100
\end{lstlisting}

On Linux, user systemd must be available. On macOS, \texttt{launchctl} must be
available in the logged-in graphical user session. Use
\texttt{run-deployment} as the foreground fallback.

\subsection*{Doctor shows warnings}

Warnings about a missing store/log before first execution or an uninstalled
service before \texttt{start} are expected. A \texttt{FAIL} result blocks safe
startup and should be corrected.

\subsection*{A changed input does not create a new finite run}

Durable stores intentionally replay prior committed work. Use a new deployment
name or explicit new \texttt{--store} for a distinct finite job. Long-running
workflows normally retain their store across restarts and redeployments.

\subsection*{The semantic diff shows an unexpected change}

Use the full and local views to locate it. Do not dismiss unexpected lifeline,
ownership, message, action-kind, control, protocol-order, or deployment changes
as formatting. The semantic diff already ignores ordinary action formatting
and comments.

\section{Safe stopping, preservation, and cleanup}

If you started the supervised tutorial deployment, stop it before archiving its
data:

\begin{lstlisting}[style=shell]
export ZIPPERGEN_HOME="$HOME/Projects/zippergen-tutorial/tutorial-runtime"
zippergen stop tutorial-review-service
\end{lstlisting}

The foreground deployment needs no stop command after it exits. To preserve all
tutorial profiles, stores, logs, environments, and bundles while keeping the
Git checkout in place, move the exact runtime directory:

\begin{lstlisting}[style=shell]
mv "$HOME/Projects/zippergen-tutorial/tutorial-runtime" \
  "$HOME/Projects/zippergen-tutorial-runtime.saved"
\end{lstlisting}

If the destination already exists, choose a different explicit name. Moving is
safer than recursive deletion and keeps the audit trail recoverable.

To reset only this project's private Studio/development context while retaining
the visible project, reopen Studio from the project root and enter:

\begin{lstlisting}[style=shell]
project reset state
\end{lstlisting}

The preview separates preserved project files from affected private state.
Answer \texttt{y} only after checking the project path and counts. On success,
Studio prints the exact backup directory below \path{ZIPPERGEN_HOME/resets} and
continues as \texttt{zippergen [no workflow]>}. Assistant requests, Studio
command history, run records, SQLite development stores, private development
keys, model/provider choices, and generated \path{.zippergen} task data are in
that backup rather than being deleted. The canonical specification, workflow
source, tests, manifest, Git
repository, and framework checkout remain where they were. Deployment
profiles, stores, logs, and running services are not part of this reset; stop
them separately as shown above. The explicit
\texttt{project reset state --yes} form is intended for scripts after
equivalent checks, not casual interactive use. To restart the design tutorial
instead, choose the fresh option from plain \texttt{project reset}; its preview
shows that the manifest, specification, and legacy prompt material will also
be archived while source, tests, and Git remain.

To abandon the workbench branch only after committing or otherwise preserving
anything important, switch back normally:

\begin{lstlisting}[style=shell]
git status --short
git switch main
\end{lstlisting}

Do not use destructive Git reset commands merely to clean up a tutorial.

\section{End-to-end acceptance checklist}

Record the observed result next to every item:

\begin{longtable}{p{0.05\linewidth}p{0.70\linewidth}p{0.16\linewidth}}
\toprule
 & \textbf{Check} & \textbf{Result/date} \\
\midrule
\endhead
$\square$ & Studio opens the tutorial workflow without a
\texttt{ZIPPERGEN\_HOME} export or explicit store path. & \\
$\square$ & \texttt{project show} reports the parent project, nested framework
checkout, and automatic \path{specification.md} path; the visible files contain
no private runtime state. & \\
$\square$ & \texttt{current}, \texttt{use}, and the selectable \texttt{show}
views make specification/refinement state, workflow name, participants, validation, model
routes, provider configuration, and runtime context clear. & \\
$\square$ & \texttt{editor set micro} remembers the installed project editor;
\texttt{editor show}, \texttt{edit file}, and a one-off
\texttt{--editor nano} behave as documented without MCP setup. & \\
$\square$ & Plain \texttt{create} opens \path{specification.md} without asking
for a filename, creates \path{.zippergen/current-task.md}, and prints a concise
result table; \texttt{task show} exposes the complete task. & \\
$\square$ & \texttt{assistant} launches Codex in the project root without a
copied path or MCP setup; the assistant creates visible source plus a focused
test at the requested paths. & \\
$\square$ & After the assistant returns, \texttt{task} reports
\texttt{awaiting human review}, states that nothing is scheduled, and lists
review commands instead of suggesting another assistant run. & \\
$\square$ & \texttt{spec refine} creates or reopens exactly one automatically
named pending refinement; the \texttt{refine} alias and repeated edits do not
create a pile of prompt files. & \\
$\square$ & The assistant receives the canonical specification plus pending
change; \texttt{spec reconcile} requires prior integration, performs no hidden
merge, archives the change, and clears the pending transaction only after
confirmation. & \\
$\square$ & \texttt{models show} lists only LLM-active lifelines, default plus
per-lifeline routes survive reopening Studio, and \texttt{models check} verifies
their effective models without changing the profile. & \\
$\square$ & Studio \texttt{run} collects both inputs and presents rejection and
approval in the same terminal. & \\
$\square$ & Studio \texttt{resume} continues the same managed run after an
interruption. & \\
$\square$ & \texttt{providers} distinguishes stored API credentials from
timestamped local checks without displaying keys; secrets remain absent from
source/request/run JSON and their private file has owner-only permissions. & \\
$\square$ & \texttt{zippergen --help} works from the project root. & \\
$\square$ & Direct mock run returns an approved and a rejected result. & \\
$\square$ & \texttt{validate} succeeds for all three projections. & \\
$\square$ & Global, communications, selected-agent, and local views are clear. & \\
$\square$ & SQLite run reaches \texttt{waiting}; CLI approval reaches
\texttt{done}. & \\
$\square$ & A pre-edit semantic snapshot is created outside the source tree. & \\
$\square$ & The assistant changes only the requested protocol behavior. & \\
$\square$ & Semantic diff matches the intended change; tests and type checks pass. & \\
$\square$ & Guided deploy creates a bundle, environment, profile, and doctor report. & \\
$\square$ & Named foreground deployment reaches a durable task and result. & \\
$\square$ & Optional supervised restart preserves the pending task. & \\
$\square$ & Logs, status, trace, doctor, restart, and stop commands are understood. & \\
$\square$ & No secret appears in source, Git diff, public profile, or service file. & \\
$\square$ & Plain \texttt{project reset} makes fresh-design, state-only, and
cancel scopes explicit. The state-only choice preserves every project file;
the fresh choice recoverably archives the manifest, specification, legacy
prompts, and private state while preserving source, tests, Git, framework
checkout, and deployment artifacts. & \\
\bottomrule
\end{longtable}

If a step is difficult or surprising, that is product feedback. Save the exact
command, output, operating system, and what you expected. The purpose of this
exercise is not only to test your understanding; it is to test whether the
development and deployment experience is genuinely self-explanatory.

\appendix

\section{Compact command sheet}
\label{app:commands}

\subsection*{Recommended Studio path}

\begin{lstlisting}[style=shell]
cd "$HOME/Projects/zippergen-tutorial"
zippergen
\end{lstlisting}

Enter the following only after the \texttt{zippergen [workflow]>} prompt
appears; do not prefix them with \texttt{zippergen}:

\begin{lstlisting}[style=shell]
project show
current
editor show
editor set micro      # use nano if micro is not installed
create                # opens the automatic specification.md
spec show
spec edit
spec path
task
task show
assistant codex      # or: assistant claude
# After the coding assistant creates source:
use                  # select the discovered workflow by number
show
validate
models reset all
models show
models check
run
resume
runs
spec refine           # creates or reopens one pending change
spec pending
assistant codex      # or: assistant claude
spec show             # verify the change was integrated
spec reconcile        # human-confirm and clear the pending change
spec history
\end{lstlisting}

The listing above stays on the free mock backend. Only after following Optional
Step~9's account, billing, and key-safety instructions, the optional paid-model
commands are:

\begin{lstlisting}[style=shell]
providers set openai
providers set anthropic
models set Writer openai:gpt-4o-mini
models set Reviewer claude:claude-sonnet-4-6
models show
models check
run
models reset all
\end{lstlisting}

Only after reading the deployment chapters and choosing mock/fake services,
enter this prepare-first operational sequence:

\begin{lstlisting}[style=shell]
models reset all
deploy tutorial-review-studio --no-start
doctor
status
start
status
logs
restart
status
stop
exit
\end{lstlisting}

No \texttt{ZIPPERGEN\_HOME} export or \texttt{STORE} variable is required for
this path. Real model overrides are optional; use \texttt{models reset all} to
stay on the free mock path. If a coding assistant is unavailable, select
\path{zippergen/examples/tutorial_review.py:tutorial_review} instead of the generated
workflow.

\subsection*{Setup: new checkout}

\begin{lstlisting}[style=shell]
mkdir -p "$HOME/Projects"
cd "$HOME/Projects"
mkdir zippergen-tutorial
cd zippergen-tutorial
git init
git clone https://github.com/zippergen-io/zippergen.git zippergen
uv python install 3.12
uv sync --project zippergen --python 3.12
uv tool install --force --editable ./zippergen
zippergen --help
zippergen studio --project .
# Inside Studio:
project init zippergen-tutorial
\end{lstlisting}

\subsection*{Inspect and validate}

\begin{lstlisting}[style=shell]
zippergen validate zippergen/examples/tutorial_review.py:tutorial_review
zippergen show zippergen/examples/tutorial_review.py:tutorial_review --detail overview
zippergen show zippergen/examples/tutorial_review.py:tutorial_review --detail protocol
zippergen show zippergen/examples/tutorial_review.py:tutorial_review \
  --detail protocol --communications
zippergen show zippergen/examples/tutorial_review.py:tutorial_review --detail actions
zippergen show zippergen/examples/tutorial_review.py:tutorial_review --detail full
zippergen show zippergen/examples/tutorial_review.py:tutorial_review --agent Reviewer
zippergen show zippergen/examples/tutorial_review.py:tutorial_review \
  --agents Writer,Reviewer
\end{lstlisting}

\subsection*{Durable local execution}

\begin{lstlisting}[style=shell]
export ZIPPERGEN_HOME="$HOME/Projects/zippergen-tutorial/tutorial-runtime"
STORE="$ZIPPERGEN_HOME/runs/manual-review.sqlite"

zippergen run zippergen/examples/tutorial_review.py:tutorial_review \
  --llm mock \
  --input 'request=Explain the sky.' --input 'max_retries=2' \
  --store "$STORE" --timeout 0

zippergen status --store "$STORE"
zippergen tasks --store "$STORE"
zippergen approve --store "$STORE" \
  --task <TASK-ID> --yes
zippergen trace --store "$STORE"
\end{lstlisting}

\subsection*{Refinement safety gate}

\begin{lstlisting}[style=shell]
export ZIPPERGEN_HOME="$HOME/Projects/zippergen-tutorial/tutorial-runtime"
mkdir -p "$ZIPPERGEN_HOME/snapshots"
SNAPSHOT="$ZIPPERGEN_HOME/snapshots/workflow-before-$(date +%Y%m%d-%H%M%S).json"
zippergen snapshot path/to/workflow.py:workflow \
  -o "$SNAPSHOT"
# edit and test the workflow
zippergen validate path/to/workflow.py:workflow
zippergen diff "$SNAPSHOT" \
  path/to/workflow.py:workflow
\end{lstlisting}

\subsection*{Deployment and operation}

\begin{lstlisting}[style=shell]
export ZIPPERGEN_HOME="$HOME/Projects/zippergen-tutorial/tutorial-runtime"
zippergen deploy zippergen/examples/tutorial_review.py:tutorial_review \
  --name tutorial-review --llm mock \
  --input 'request=Explain deployment.' --input 'max_retries=2' \
  --yes --no-start
zippergen doctor tutorial-review
# Recommended first execution; approve it from a second terminal:
zippergen run-deployment tutorial-review
# Optional supervised-service exercise:
zippergen start tutorial-review --dry-run
zippergen start tutorial-review --enable
zippergen status tutorial-review
# Press Control-C when you have seen enough output; this does not stop service:
zippergen logs tutorial-review --follow
zippergen restart tutorial-review
zippergen configure tutorial-review --yes --restart
zippergen deploy tutorial-review --yes
zippergen stop tutorial-review
\end{lstlisting}

\section{Complete tutorial source}
\label{app:source}

The listing below is imported from the executable file at document build time.
Compile this document from the nested framework checkout so the relative path
resolves.

\lstinputlisting[style=python,caption={examples/tutorial\_review.py}]{examples/tutorial_review.py}

\section{Build this document}

From the tutorial project root, with a TeX distribution installed:

\begin{lstlisting}[style=shell]
cd zippergen
mkdir -p docs/_build
latexmk -pdf -interaction=nonstopmode -halt-on-error \
  -outdir=docs/_build \
  docs/workflow-development-deployment-guide.tex
\end{lstlisting}

The PDF is then:
\path{zippergen/docs/_build/workflow-development-deployment-guide.pdf}. TeX auxiliary
files and the generated PDF live under the ignored \path{docs/_build}
directory; the maintained source is the \texttt{.tex} file.

\end{document}
